---
title: "Temporally Dependent Data and Methods"
subtitle: "An Introduction"
author:
  - name: "Send comments to: Tony T (tthrall)"
date: last-modified
date-format: HH:mm ddd D-MMM-YYYY
output:
  html_document:
    toc: true
    df_print: paged
    mathjax: default
  word_document:
    toc: true
    df_print: tibble
  pdf_document:
    toc: true
    df_print: tibble
abstract:
  "Introduce data examples in which successive observations are statistically dependent.  Discuss corresponding data analysis methods.  <br><br>"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  error   = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r cran_libraries}
library(assertthat)
library(astsa)
library(here)
library(knitr)
library(latex2exp)
library(tidyverse)
library(tinytex)
```

```{r local_libraries}
# library(ugrid)
```

```{r local_source}
# gen_ds_tbl
# compile a tibble that identifies data sets
source(here("code", "gen_ds_tbl.R"))

# get_ref_materials
# enscapsulate reference materials as separate modules
source(here("code", "get_ref_materials.R"))
```

```{r plotting_params}
# grid-points in (-m pi, m pi), m = m_periods
# x_vec <- u_vec * (m_periods * pi)
```

## Introduction

This technical note gives examples of data in which successive observations are statistically dependent and are typically indexed by time. Time series data sampled at regular intervals (daily stock prices, for example) are the most common case, followed by point process data representing the occurrence of events (earthquakes, for example) at random times.

Statistical models of such data include parametric models expressed in the time domain and non-parametric models expressed in the Fourier frequency domain (number of cycles per unit of time).

This note also comments on the detection of statistical dependence among successive observations and the modification of statistical tests and estimators for such data.

## Examples

```{r astsa_data_set_tbl}
astsa_data_set_tbl <- get_all_ds_per_pkg(pkgs = "astsa")

# write to file only as needed
save_to_file <- FALSE
if (save_to_file) {
  # TSV
  astsa_data_set_tbl |>
    readr::write_tsv(here(
	  "data", "retain", "astsa_data_set_tbl.txt"
	))
  # RDS
  astsa_data_set_tbl |>
    readr::write_rds(here(
	  "data", "rds", "astsa_data_set_tbl.rds"
	))
}
```

### EXAMPLE_TOPIC

#### Description

#### Questions and Objectives

### Climate Change

```{r climate_ds_tbl}
# names of relevant data sets from package astsa::
climate_ds_names <- c(
  "ENSO", "gtemp.month", "gtemp_both",
  "gtemp_land", "gtemp_ocean", "soi"
)
climate_ds_tbl <- select_ds_per_pkg(
  pkgs = "astsa",
  datasets = climate_ds_names)
```

#### Description

The `R` package `astsa` provides the following data sets related to the analysis of climate change.

```{r climate_ds_tbl__kable}
climate_ds_tbl |> knitr::kable(
  caption = "Climate: selected data sets"
)
```

#### Questions and Objectives

## Data Analysis Objectives

## Data Analysis Methods

## Detecting Observational Dependence

## Closing Remarks

## Glossary of Math Symbols

```{r symbol_tbl}
symbol_tbl <- get_symbol_tbl()
```

```{r symbol_tbl__save}
# save file only when updated
save_to_file <- FALSE
if (save_to_file) {
  # RDS format
  symbol_tbl |> readr::write_rds(here::here(
    "data", "retain", "symbol_tbl.rds"
  ))

  # TSV format
  symbol_tbl |> readr::write_tsv(here::here(
    "data", "retain", "symbol_tbl.txt"
  ))
}
```

```{r symbol_tbl__kable}
symbol_tbl |>
  dplyr::select(symbol, dscr) |>
  knitr::kable(
    format = "html",
	escape = FALSE,
	caption = "Glossary of Mathematical Symbols (s)",
	col.names = c("s", "description")
  )
```

## Resources

### Books

```{r books_tbl}
books_tbl <- get_books_tbl() |>
  dplyr::arrange(title)
```

```{r books_tbl__save}
# save file only when updated
save_to_file <- FALSE
if (save_to_file) {
  # RDS format
  books_tbl |> readr::write_rds(here::here(
    "data", "retain", "books_tbl.rds"
  ))

  # TSV format
  books_tbl |> readr::write_tsv(here::here(
    "data", "retain", "books_tbl.txt"
  ))
}
```

```{r books_tbl__kable}
books_tbl |>
  list_links(a_idx = 3L) |>
  knitr::kable(
    format    = "html",
	escape    = FALSE,
	col.names = ""
  )
```

### Articles

```{r articles_tbl}
articles_tbl <- get_articles_tbl() |>
  dplyr::arrange(title)
```

```{r articles_tbl__save}
# save file only when updated
save_to_file <- FALSE
if (save_to_file) {
  # RDS format
  articles_tbl |> readr::write_rds(here::here(
    "data", "retain", "articles_tbl.rds"
  ))

  # TSV format
  articles_tbl |> readr::write_tsv(here::here(
    "data", "retain", "articles_tbl.txt"
  ))
}
```

```{r articles_tbl__kable}
articles_tbl |>
  list_links() |>
  knitr::kable(
    format    = "html",
	escape    = FALSE,
	col.names = ""
  )
```
