---
title: "Latent Dirichlet Allocation"
subtitle: "Selected topics from Part 1 of Data Mining Intro"
author: 
  - name: "Send comments to: Tony T (tthrall)"
date: last-modified
date-format: HH:mm ddd D-MMM-YYYY
editor: source
toc: true
toc-depth: 3
format:
  html:
    toc_float: true
    code-fold: true
    code-summary: "Show the code"
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
  pdf: default
  docx: default
params:
  file_id: "xtra_latent_dir_alloc"
  file_prefix: "xtra_latent_dir_alloc"
abstract: 
  "Introduce latent Dirichlet allocation (LDA) using Gibbs sampling.  <br><br>"
bibliography: xtra_latent_dir_alloc.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo    = FALSE, 
  error   = FALSE, 
  message = FALSE, 
  warning = FALSE
)
```

```{r}
#| label: CRAN-libraries

# library(assertthat)
# library(GGally)
# library(gutenbergr)
library(here)
# library(ISLR2)
# library(janeaustenr)
# library(latex2exp)
# library(lda)
# library(quanteda)
library(tidytext)
library(tidyverse)
library(tinytex)
library(tm)
library(tokenizers)
# library(topicmodels)
# library(tufte)

# package topic models available?
topicmodels_loaded <- require(topicmodels)

```

```{r}
#| label: local-source

source(here("code", "text_helpers.R"))
```

------------------------------------------------------------------------

## Background

Latent Dirichlet Allocation (LDA) was proposed by @BNJ2003LDA as a method of topic modeling. The method is briefly mentioned in Part 1 of the course. Several course participants requested a more detailed description. This note is a response to that request.

## Text Analysis

Part 1, Session 2b, of the course presents an overview of text analysis based on @Silge_Robinson_2017.  The authors use the `R` function `topicmodels::LDA()` with default parameter setting `method = VEM`, which denotes the Variational Expectation Maximization (EM) algorithm.  In this note we use an alternative implementation of LDA described below.

## Generative Model

LDA treats each document as a mixture of topics, and each topic as a mixture of words.  LDA adopts a _bag-of-words_ model in which each document $d$ in a corpus of documents $\mathcal{D}$ is generated from two Dirichlet distributions: $Dir(\alpha_\bullet)$ over $K$ possible topics; and $Dir(\eta_\bullet)$ over $N$ possible words.  In this model documents are presumed to have been randomly generated as follows.[^generate-v-sample]

[^generate-v-sample]: Other authors use the verb "sample" rather than "generate" to mean random generation from a specified probability distribution.

  - Generate word probability vector $\phi_\bullet (k) \sim Dir(\eta_\bullet )$ for each topic index $k \in \{ 1, 2, \ldots, K \}$

Then for each document $d \in \mathcal{D}$ and for each word position $\nu \in d$:[^doc-length]

[^doc-length]:  We assume that document $d$ consists of $n_d$ word positions, which is either a fixed positive integer, or else is generated independently of other random variables.

  1. Generate topic probability vector $\theta_\bullet (d) \sim Dir(\alpha_\bullet)$
  1. Assign topic index $\kappa (d,\nu) \sim \theta_\bullet (d)$
  1. Generate word index $\omega (d,\nu) \sim \phi_\bullet (\kappa (d,\nu))$

These Dirichlet distributions are typically assumed to be symmetric.  That is, if we define 

$$
\begin{align}
  \bar{\alpha} &= \frac{\alpha_+}{K} = \frac{1}{K} \sum_{k = 1}^K
\alpha_k \\ 
  \bar{\eta} &= \frac{\eta_+}{N} = \frac{1}{N} \sum_{\nu = 1}^N
\eta_\nu
\end{align}
$$ {#eq-avg-Dirichlet-param}

then current practice is to set all elements of vectors $\alpha_\bullet$ and $\eta_\bullet$ equal to their respective average values.

$$
\begin{align}
  \alpha_k &= \bar{\alpha} \quad \text{for } k \in \{1, \ldots, K \} \\ 
  \eta_\nu &= \bar{\eta} \quad \text{for } \nu \in \{1, \ldots, N \}
\end{align}
$$ {#eq-symm-Dirichlet}

The values $\bar{\alpha}$ and $\bar{\eta}$ are called _concentration_ parameters.

For a symmetric Dirichlet distribution as above, if $\bar{\alpha} < 1$ then for each document $d$, $Dir(\alpha_\bullet)$ tends to generate a sparse probability vector $\theta_\bullet (d)$ that gives most of its weight to just a few of the K topics (with the selection of topics varying across documents).  On the other hand, if $\bar{\alpha} > 1$ then $Dir(\alpha_\bullet)$ tends to generate a more uniform probability vector $\theta_\bullet (d)$ spread more evenly across all topics.

As for the distribution of words within a topic, in most applications the number $N$ of distinct words in the corpus $\mathcal{D}$ may be a few thousand or even a few tens of thousands.  Then each topic is a mixture across a distinct subset of a large number of words, typically with $\bar{\eta} < 1$.

See @Antoniak_2023 for practical suggestions on these and other settings and workflows.

## Book Review Example

### Data

"Animal Farm" by George Orwell was published in 1945, and nearly forty years later (1984) "The Butter Battle Book" by Dr. Seuss was published.  The books are quite different of course, but they both allude to the Soviet Union in its earlier and later years, respectively.

In 1945 "Animal Farm" by George Orwell was published, and nearly forty years later (1984) "The Butter Battle Book" by Dr. Seuss was published.  The books are quite different of course, but they both allude to the Soviet Union in its earlier and later years, respectively.

As a toy example of a corpus of documents, we have extracted Wikipedia's description of the plots of the two books, counting each paragraph as a separate document.

```{r}
#| label: af-pst-lst

# Animal Farm: story plot from Wikipedia
# list of tibbles: paragraphs, sentences, tokens
af_pst_lst <- para2token(
  file = here::here(
    "data", "retain", "Animal_Farm_summary.txt")
)

af_para_tbl <- af_pst_lst$ para_tbl |> 
  dplyr::mutate(src = "AF") |> 
  dplyr::select(src, everything())

af_token_tbl <- af_pst_lst$ token_tbl |> 
  dplyr::mutate(src = "AF") |> 
  dplyr::select(src, everything())
```

```{r}
#| label: bbb-pst-lst

# The Butter Battle Book: story plot from Wikipedia
# list of tibbles: paragraphs, sentences, tokens
bbb_pst_lst <- para2token(
  file = here::here(
    "data", "retain", "Butter_Battle_summary.txt")
)

bbb_para_tbl <- bbb_pst_lst$ para_tbl |> 
  dplyr::mutate(src = "BBB") |> 
  dplyr::select(src, everything())

bbb_token_tbl <- bbb_pst_lst$ token_tbl |> 
  dplyr::mutate(src = "BBB") |> 
  dplyr::select(src, everything())
```

```{r}
#| label: af-bbb-para-tbl

af_bbb_para_tbl <- dplyr::bind_rows(
  af_para_tbl, 
  bbb_para_tbl
)
```

Here are the successive paragraphs (stripped of punctuation and stop-words) summarizing the two books.

```{r}
#| label: tbl-af-bbb-para-tbl
#| tbl-cap: "Paragraphs from summaries of two books"
#| echo: true

af_bbb_para_tbl |> print(width = 72)
```


```{r}
#| label: af-bbb-token-tbl

af_bbb_token_tbl <- dplyr::bind_rows(
  af_token_tbl, 
  bbb_token_tbl
)
```

From these paragraphs, each treated as a document, we create the following `DocumentTermMatrix` object, which is a list that includes: 

  - a sparse matrix giving the document-index, the term-index, and the frequency (number of occurrences) of the indexed term; and 
  - a vector of the terms themselves

```{r}
#| label: af-bbb-dtm-lst
#| echo: true

af_bbb_dtm_lst <- 
  tm::DocumentTermMatrix(x = af_bbb_para_tbl$ para)

# af_bbb_dtm_lst  |> object.size(): 44 KB
# 44464 bytes

af_bbb_dtm_lst
```

Here is a table showing the frequency of each term within each document (paragraph).

```{r}
#| label: tbl-af-bbb-tf-tbl
#| tbl-cap: "Term-frequency: number of occurrences of each term within each document"
#| echo: true

af_bbb_tf_tbl <- af_bbb_dtm_lst |> tidytext::tidy()
af_bbb_tf_tbl
```

### LDA function

We now employ function `topicmodels::LDA()`[^art-Grün_Hornik_2011] to perform latent Dirichlet Allocation.  The function determines the probable topic of each word within each document by one of two methods: (1) "VEM" (Variational Expectation Maximization); or else (2) "Gibbs" (Gibbs Sampling).  The function output, depending on the choice of method, is a list conforming to the format of either `LDA_VEM` or else `LDA_Gibbs`.  Here we choose the Gibbs Sampling method because it requires less memory.

[^art-Grün_Hornik_2011]:  See @Grün_Hornik_2011.

```{r}
#| label: lst-af-bbb-gs
#| lst-cap: ""
#| echo: true
#| eval: false

# Latent Dirichlet Allocation via Gibbs Sampling

extract_from_DTM <- FALSE
if (extract_from_DTM && topicmodels_loaded) {
  af_bbb_gs <- af_bbb_dtm_lst |> 
    topicmodels::LDA(
      # number of topics
      k = 2, 
      method = "Gibbs", 
      # set seed to ensure results can be reproduced
      control = list(seed = 1234)
    )
} else {
  # TODO: decide whether and how to store LDA_Gibbs output
}

# af_bbb_gs |> object.size(): 77 KB
# 77032 bytes
```

From the `LDA_Gibbs` output we extract the following "beta" matrix, that is the matrix of fitted probabilities that a term within a given document belongs to topic 1 or 2, respectively.

```{r}
#| label: af-bbb-topics
#| eval: true
#| echo: true

# TODO: overcome error generated by tidytext::tidy(<LDA_Gibbs>)
# when <LDA_Gibbs> is large and memory is constrained
# in the meantime, patch as follows

extract_from_DTM <- FALSE
if (extract_from_DTM) {
  # beta: K (topics) by N (terms) of log-probabilities
  beta_mat           <- af_bbb_gs@beta
  colnames(beta_mat) <- af_bbb_gs@terms
  rownames(beta_mat) <- paste0("topic_", 1:2)
  
  beta_tbl           <- beta_mat |> 
    tibble::as_tibble(rownames = "topic") |> 
    dplyr::mutate(topic = 1:2)
  
  beta_long          <- beta_tbl |> 
    tidyr::pivot_longer(
      cols      = - topic, 
      names_to  = "term", 
      values_to = "ln_prob"
    ) |> 
    dplyr::arrange(term)
  
  # af_bbb_topics: N by K tibble of probabilities
  af_bbb_topics <- beta_long |> 
    dplyr::mutate(
      beta = exp(ln_prob)
    ) |> 
    dplyr::select(- ln_prob)
  
  # save af_bbb_topics as TSV file only as needed
  save_file <- FALSE
  if (save_file) {
    af_bbb_topics |> readr::write_tsv(here::here(
      "data", "retain", "af_bbb_topics.txt"
    ))
  }
  
} else {
  # read in previously saved file
  af_bbb_topics <- readr::read_tsv(here::here(
    "data", "retain", "af_bbb_topics.txt"
  ))
}

# af_bbb_topics |> object.size(): 49 KB
# 48992 bytes

af_bbb_topics |> print(n = 4, digits = 2)
```


### Output interpretation

Here are the most probable terms for each of the two constructed topics, along with their probabilities (beta), shown as a bar chart.

```{r}
#| label: af-bbb-top-terms
#| eval: true

# set with_ties = F to limit returned tibble to <= n rows
af_bbb_top_terms <- af_bbb_topics |> 
  dplyr::slice_max(beta, n = 10, by = "topic", with_ties = FALSE)
```

```{r}
#| label: fig-af-bbb-top-terms
#| fig-cap: "AF-BBB paragraphs: most probable terms by topic"
#| echo: true

g_af_bbb_top_terms <- af_bbb_top_terms |> 
  dplyr::mutate(term = reorder_within(
    x = term, 
    by = beta, 
    within = topic)) |> 
  ggplot2::ggplot(mapping = aes(
    x = beta, y = term, fill = factor(topic)
  )) + 
  ggplot2::geom_col(show.legend = FALSE) + 
  ggplot2::facet_wrap(~ as_factor(topic), scales = "free") + 
  tidytext::scale_y_reordered() + 
  ggplot2::labs(title = "AF-BBB paragraphs: most probable terms by topic")
g_af_bbb_top_terms
```

Topics 1 and 2 seem to pertain to "The Butter Battle Book" and "Animal Farm", respectively, although we see "farm" in topic 1 and "battle"  in topic 2.

Another way to compare topics 1 and 2 is to examine the terms shared by the two topics and then find the terms having the biggest disparity in (topic, term) probability (beta). Here's a bar chart showing the more prominent differences, expressed as

$$
\begin{align}
  \log_2 \left( \frac{\beta_2}{\beta_1} \right)
\end{align}
$$ {#eq-log2-ratio-topic-term}

restricting the set of terms to those assigned to both topics $(\min(\beta_1, \beta_2) > 0)$ with at least one of them exceeding a probability threshhold, say $(\max(\beta_1, \beta_2) > 0.001)$.

```{r}
#| label: af-bbb-b-ratio-wide

af_bbb_b_ratio_wide <- af_bbb_topics |> 
  dplyr::mutate(topic = paste0("beta_", topic)) |> 
  tidyr::pivot_wider(names_from = topic, values_from = beta) |> 
  dplyr::filter(
    pmin(beta_1, beta_2) > 0, 
    pmax(beta_1, beta_2) > 0.001
  ) |> 
  dplyr::mutate(log_ratio = log2(beta_2 / beta_1))
```

```{r}
#| label: fig-af-bbb-b-ratio-wide
#| fig-cap: "Log2 ratio of beta in topic 2 / topic 1"
#| echo: true

g_af_bbb_b_ratio_wide <- af_bbb_b_ratio_wide |> 
  dplyr::group_by(direction = log_ratio > 0) |> 
  dplyr::slice_max(abs(log_ratio), n = 10) |> 
  dplyr::ungroup() |> 
  dplyr::mutate(term = reorder(term, log_ratio)) |> 
  ggplot2::ggplot(aes(log_ratio, term)) +
  ggplot2::geom_col() +
  ggplot2::labs(x = "Log2 ratio of beta in topic 2 / topic 1", y = NULL)
g_af_bbb_b_ratio_wide
```

This figure further supports the earlier conjecture that topic 1 best matches "The Butter Battle Book" and topic 2 best matches "Animal Farm".

## The LDA-Gibbs Algorithm

The LDA identification of topics is a particular form of Markov chain Monte Carlo (MCMC) method.  The data can be summarized as the set of unique words across the corpus of documents, along with counts of those unique words per document.  The objective is to determine the topic of each word instance within each document.

Assignment of topics to word positions is intially made at random in a manner similar to the generative model outlined above.  We thereby have initial values for: 

$$
\begin{align}
  n_{d, k} &= \text{number of assignments of topic } k \text{ to document } d \\ 
  n_{w, k} &= \text{number of assignments of topic } k \text{ to unique word } w \\ 
  n_k &= \text{number of assignments of topic } k \text{ across corpus } \mathcal{D} 
\end{align}
$$ {#eq-log-ratio-topic-term}

Then through many iterations the prior distributions $Dir(\alpha_\bullet)$ and $Dir(\eta_\bullet)$ are updated to form posterior distributions based on the data.

The Gibbs sampling procedure treats each word position in the text collection in turn, and estimates the probability of assigning the word in that position to each topic, conditioned on the current topic assignments of all other words.  From this conditional distribution, a topic is generated and stored as the new topic assignment for the given word in the given position.

A great computational savings is achieved by the choice of a Dirichlet prior for the multinomial distribution of unique topic instances per document, because the Dirichlet distribution is conjugate to the multinomial distribution.  For further details on this updating algorithm see @Steyvers_Griffiths_2007.

## Concluding Remarks

Topic modeling is an active area of research that contributes to the statistical analysis of large document collections.  The LDA model of document-generation enables Bayesian analysis of the latent structure of a corpus of documents.

## References
