---
title: "Latent Dirichlet Allocation"
subtitle: "Selected topics from Part 1 of Data Mining Intro"
author: 
  - name: "Send comments to: Tony T (tthrall)"
date: last-modified
date-format: HH:mm ddd D-MMM-YYYY
editor: source
toc: true
toc-depth: 3
format:
  html:
    toc_float: true
    code-fold: true
    code-summary: "Show the code"
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
  pdf: default
  docx: default
params:
  file_id: "xtra_latent_dir_alloc"
  file_prefix: "xtra_latent_dir_alloc"
abstract: 
  "Introduce latent Dirichlet allocation (LDA) using Gibbs sampling.  <br><br>"
bibliography: xtra_latent_dir_alloc.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo    = FALSE, 
  error   = FALSE, 
  message = FALSE, 
  warning = FALSE
)
```

```{r}
#| label: CRAN-libraries

library(assertthat)
library(GGally)
library(gutenbergr)
library(here)
library(ISLR2)
library(janeaustenr)
library(latex2exp)
# library(lda)
library(quanteda)
library(tidytext)
library(tidyverse)
library(tinytex)
library(tm)
library(tokenizers)
# library(topicmodels)
library(tufte)

# package topic models available?
topicmodels_loaded <- require(topicmodels)

```

```{r}
#| label: local-source

source(here("code", "rmse_per_grp.R"))
source(here("code", "xtabs_to_jaccard.R"))
```

------------------------------------------------------------------------

## Background

Latent Dirichlet Allocation (LDA) was proposed by @BNJ2003LDA as a method of topic modeling. The method is briefly mentioned in Part 1 of the course. Several course participants requested a more detailed description. This note is a response to that request.

## Text Analysis

Part 1, Session 2b, of the course presents an overview of text analysis based on @Silge_Robinson_2017.  The authors use the `R` function `topicmodels::LDA()` with default parameter setting `method = VEM`, which denotes the Variational Expectation Maximization (EM) algorithm.  In this note we use an alternative implementation of LDA described below.

## Document Generating Model

LDA treats each document as a mixture of topics, and each topic as a mixture of words.  LDA assumes that each document $d$ in a corpus of documents $\mathcal{D}$ is generated from two Dirichlet distributions: $Dir(\alpha_\bullet)$ over $K$ possible topics; and $Dir(\beta_\bullet)$ over $N$ possible words.  This is a so-called bag-of-words model of the following form.

  - Generate word probability vector $\phi_\bullet (k) \sim Dir(\beta_\bullet )$ for each topic index $k \in \{ 1, 2, \ldots, K \}$

Then for each document $d \in \mathcal{D}$ and for each word position $\nu \in d$:

  1. Generate topic probability vector $\theta_\bullet (d) \sim Dir(\alpha_\bullet)$
  1. Assign topic index $\kappa (d,\nu) \sim \theta_\bullet (d)$
  1. Generate word index $\omega (d,\nu) \sim \phi_\bullet (\kappa (d,\nu))$

## News Article Example

### Data

The `R` package `topicmodels` includes the `AssociatedPress` data representing 2246 news articles from the AP news agency, mostly published around 1988.  The data are in the form of a list[^DTM] that includes a sparse matrix giving the document-index, the term-index, and the frequency (number of occurrences) of the indexed term (word) within each document (news article).  The list also contains a vector of the terms themselves.

[^DTM]: The Document Term Matrix (DTM) is a list format created by `R` package `tm`, devoted to topic mining.

### LDA function

The `topicmodels` package also includes function `LDA()` that performs latent Dirichlet Allocation.  The function determines the probable topic of each word within each document by one of two methods: (1) "VEM" (Variational Expectation Maximization); or else (2) "Gibbs" (Gibbs Sampling).  The function output, depending on the choice of method, is a list conforming to the format of either `LDA_VEM` or else `LDA_Gibbs`.  Here we choose the Gibbs Sampling method because it requires less memory.

```{r}
#| label: lst-Associated-Press
#| lst-cap: ""
#| echo: true
#| eval: false

# avoid `topicmodels` if possible: it's slow to load
if (topicmodels_loaded) {
  data("AssociatedPress", package = "topicmodels")
}

# AssociatedPress |> object.size(): 5.5 MB

# AssociatedPress
# <<DocumentTermMatrix (documents: 2246, terms: 10473)>>
# Non-/sparse entries: 302031/23220327
# Sparsity           : 99%
# Maximal term length: 18
# Weighting          : term frequency (tf)

# str(AssociatedPress)
# List of 6
#  $ i       : int [1:302031] 1 1 1 1 1 1 1 1 1 1 ...
#  $ j       : int [1:302031] 116 153 218 272 299 302 447 455 548 597 ...
#  $ v       : num [1:302031] 1 2 1 1 1 1 2 1 1 1 ...
#  $ nrow    : int 2246
#  $ ncol    : int 10473
#  $ dimnames:List of 2
#   ..$ Docs : NULL
#   ..$ Terms: chr [1:10473] "aaron" "abandon" "abandoned" "abandoning" ...
#  - attr(*, "class")= chr [1:2] "DocumentTermMatrix" "simple_triplet_matrix"
#  - attr(*, "weighting")= chr [1:2] "term frequency" "tf"

# object.size(AssociatedPress)
# 5539768 bytes
```

Here are the first few terms extracted from the news articles.  They are listed alphabetically and indexed in that order.

```{r}
#| label: ap-terms-tbl
#| eval: true

# extract terms from `AssociatedPress` DTM
# save as TSV file

extract_from_AP_DTM <- FALSE
if (extract_from_AP_DTM) {
  ap_terms_tbl <- tibble::tibble(
    term = AssociatedPress$ dimnames$ Terms, 
    idx  = 1:length(term)
  ) |> 
    dplyr::select(idx, term)
  
  # save as TSV file only as needed
  save_file <- FALSE
  if (save_file) {
    ap_terms_tbl |> readr::write_tsv(here::here(
      "data", "retain", "ap_terms_tbl.txt"
    ))
  }
} else {
  ap_terms_tbl <- readr::read_tsv(here::here(
    "data", "retain", "ap_terms_tbl.txt"
  ))
}

# ap_terms_tbl |> object.size(): 0.75 MB
# 748208 bytes

ap_terms_tbl |> print(n = 4)
```

Here is the document-term matrix in sparse format.  The columns denote the index of the document, the index of the term, and the term-frequency, that is, the number of occurrences of the term within the document.

```{r}
#| label: ap-dtm-tbl
#| eval: true

# extract term-counts per doc from `AssociatedPress` DTM
# save as TSV file

extract_from_AP_DTM <- FALSE
if (extract_from_AP_DTM) {
  # document term matrix (DTM): sparse format
  # the actual terms are stored in ap_terms_tbl
  ap_dtm_tbl <- tibble::tibble(
    # document (row) index
    doc_idx = AssociatedPress$ i, 
    # term (column) index
    trm_idx = AssociatedPress$ j, 
    # term-frequency count
    tf_ct   = AssociatedPress$ v
  )
  
  # save as TSV file only as needed
  save_file <- FALSE
  if (save_file) {
    ap_dtm_tbl |> readr::write_tsv(here::here(
      "data", "retain", "ap_dtm_tbl.txt"
    ))
  }
} else {
  ap_dtm_tbl <- readr::read_tsv(here::here(
    "data", "retain", "ap_dtm_tbl.txt"
  ))
}

# ap_dtm_tbl |> object.size(): 7.3 MB
# 7252896 bytes

ap_dtm_tbl |> print(n = 4)
```

```{r}
#| label: lst-ap-gs
#| lst-cap: ""
#| echo: true
#| eval: false

# Latent Dirichlet Allocation via Gibbs Sampling

extract_from_AP_DTM <- FALSE
if (extract_from_AP_DTM && topicmodels_loaded) {
  ap_gs <- AssociatedPress |> 
    topicmodels::LDA(
      # number of topics
      k = 2, 
      method = "Gibbs", 
      # set seed to ensure results can be reproduced
      control = list(seed = 1234)
    )
} else {
  # TODO: decide whether and how to store ap_gs
}

# ap_gs |> object.size(): 7.5 MB
# 7493512 bytes
```

From the `LDA_Gibbs` output we extract the following "beta" matrix, that is the matrix of fitted probabilities that a term within a given document belongs to topic 1 or 2, respectively.

```{r}
#| label: ap-topics
#| eval: true

# TODO: overcome error generated by tidy(ap_gs)
# in the meantime, patch as follows

extract_from_AP_DTM <- FALSE
if (extract_from_AP_DTM) {
  # beta: K (topics) by N (terms) of log-probabilities
  beta_mat           <- ap_gs@beta
  colnames(beta_mat) <- ap_gs@terms
  rownames(beta_mat) <- paste0("topic_", 1:2)
  
  beta_tbl           <- beta_mat |> 
    as_tibble(rownames = "topic") |> 
    dplyr::mutate(topic = 1:2)
  
  beta_long          <- beta_tbl |> 
    tidyr::pivot_longer(
      cols      = - topic, 
      names_to  = "term", 
      values_to = "ln_prob"
    ) |> 
    dplyr::arrange(term)
  
  # ap_topics: N by K tibble of probabilities
  ap_topics <- beta_long |> 
    dplyr::mutate(
      beta = exp(ln_prob)
    ) |> 
    dplyr::select(- ln_prob)
  
  # save ap_topics as TSV file only as needed
  save_file <- FALSE
  if (save_file) {
    ap_topics |> readr::write_tsv(here::here(
      "data", "retain", "ap_topics.txt"
    ))
  }
  
} else {
  # read in previously saved file
  ap_topics <- read_tsv(here(
    "data", "retain", "ap_topics.txt"
  ))
}

ap_topics |> print(n = 4, digits = 2)
```

### Output interpretation

Here are the most probable terms for each of the two constructed topics, along with their probabilities (beta), shown as a bar chart.

```{r}
#| label: ap-top-terms

ap_top_terms <- ap_topics |> 
  slice_max(beta, n = 10, by = "topic")
```

```{r}
#| label: ap-top-terms-wide

# afterthought: display bar chart, not this table
ap_top_terms_1 <- ap_top_terms |> 
  dplyr::filter(topic == 1L)

ap_top_terms_2 <- ap_top_terms |> 
  dplyr::filter(topic == 2L)

ap_top_terms_wide <- tibble::tibble(
  rank    = 1:nrow(ap_top_terms_1), 
  topic_1 = ap_top_terms_1$term, 
  topic_2 = ap_top_terms_2$term
)
```

```{r}
#| label: fig-ap-top-terms
#| fig-cap: "AP articles: most probable terms by topic"

g_ap_top_terms <- ap_top_terms |> 
  mutate(term = reorder_within(
    x = term, 
    by = beta, 
    within = topic)) |> 
  ggplot(mapping = aes(
    x = beta, y = term, fill = factor(topic)
  )) + 
  geom_col(show.legend = FALSE) + 
  facet_wrap(~ as_factor(topic), scales = "free") + 
  scale_y_reordered() + 
  labs(title = "AP articles: most probable terms by topic")
g_ap_top_terms
```

Topics 1 and 2 seem to pertain to business and politics, respectively, although "new" and "people" are prominent terms for both topics.

Another way to compare topics 1 and 2 is to examine the terms shared by the two topics and then find the terms having the biggest disparity in (topic, term) probability (beta). Here's a bar chart showing the more prominent differences, expressed as

$$
\begin{align}
  \log_2 \left( \frac{\beta_2}{\beta_1} \right)
\end{align}
$$ {#eq-log-ratio-topic-term}

restricting the set of terms to those assigned to both topics $(\min(\beta_1, \beta_2) > 0)$ with at least one of them exceeding a probability threshhold, say $(\max(\beta_1, \beta_2) > 0.001)$.

```{r}
#| label: beta-ratio-wide

beta_ratio_wide <- ap_topics |> 
  mutate(topic = paste0("beta_", topic)) |> 
  pivot_wider(names_from = topic, values_from = beta) |> 
  filter(
    pmin(beta_1, beta_2) > 0, 
    pmax(beta_1, beta_2) > 0.001
  ) |> 
  mutate(log_ratio = log2(beta_2 / beta_1))
```

```{r}
#| label: fig-beta-ratio-wide
#| fig-cap: "Log2 ratio of beta in topic 2 / topic 1"

g_beta_ratio_wide <- beta_ratio_wide |> 
  group_by(direction = log_ratio > 0) |> 
  slice_max(abs(log_ratio), n = 10) |> 
  ungroup() |> 
  mutate(term = reorder(term, log_ratio)) |> 
  ggplot(aes(log_ratio, term)) +
  geom_col() +
  labs(x = "Log2 ratio of beta in topic 2 / topic 1", y = NULL)
g_beta_ratio_wide
```

This figure further supports the earlier conjecture that topics 1 and 2 pertain to business and politics, respectively.

## The LDA-Gibbs algorithm

## The Conjugacy Advantage

The beauty lies in the conjugate relationships. Since Dir is conjugate to Multinomial:

**For topic proportions:** If θ_d ~ Dir(α) and we observe topic assignments z_d, then θ_d | z_d ~ Dir(α + n_d), where n_d is the vector of topic counts in document d.

**For word distributions:** If φ_k ~ Dir(β) and we observe words assigned to topic k, then φ_k | words ~ Dir(β + n_k), where n_k counts how often each word was assigned to topic k.

## Gibbs Sampling Implementation

The algorithm alternates between sampling topic assignments and updating the distributions:

**Sampling z_{d,n}:** For each word token, sample its topic assignment from: P(z_{d,n} = k | rest) ∝ (n_{d,k}^{-n} + α_k) × (n_{k,w}^{-n} + β_w) / (n_k^{-n} + Σβ)

Where the superscript -n means "excluding the current token."

**Updating parameters:** After each sampling sweep, the Dirichlet posteriors give us:

- θ_{d,k} = (n_{d,k} + α_k) / (Σ_j n_{d,j} + Σ_j α_j)
- φ_{k,w} = (n_{k,w} + β_w) / (Σ_v n_{k,v} + Σ_v β_v)

## Why This Works

The conjugacy means we can integrate out θ and φ analytically, leaving us to sample only the discrete topic assignments z. The sufficient statistics (count matrices n) fully capture everything we need to know about the posterior distributions.

The algorithm converges because it's doing collapsed Gibbs sampling on the joint posterior P(z, θ, φ | w, α, β), which has a unique stationary distribution under mild regularity conditions.

This mathematical elegance - leveraging conjugacy to avoid explicitly storing continuous parameters - is what makes LDA tractable for large document collections.

Claude: Is there anything else about LDA or related topics you'd like to explore while we're on the subject?

TT: Yes, please tell me about the number of topics (say K) used by the LDA algorithm. Can the user restrict the value of K?

Claude: 
Yes, the user must specify K (the number of topics) as a hyperparameter before running LDA. This is both a requirement and a limitation of the algorithm.

## K as a Required Input

Unlike some clustering algorithms that can determine the optimal number of clusters automatically, LDA requires you to decide on K upfront. The algorithm then finds the best K-topic solution given your choice, but it won't tell you whether K=5 or K=10 would be better for your data.

## Choosing K in Practice

**Too few topics (K too small):** Topics become overly broad and less meaningful. You might get a generic "science" topic that mixes physics, biology, and chemistry when these should be separate themes.

**Too many topics (K too large):** Topics become overly specific or start splitting coherent themes. You might end up with separate topics for "baseball games" and "baseball statistics" when they could reasonably be one "baseball" topic.

**Common approaches for selecting K:**

- Domain expertise (how many distinct themes do you expect?)
- Grid search with evaluation metrics
- Perplexity curves (though lower perplexity doesn't always mean better interpretability)
- Topic coherence measures
- Human evaluation of topic quality

## Evaluation Methods

**Perplexity** measures how well the model predicts held-out documents (lower is better), but often doesn't correlate with human judgments of topic quality.

**Topic coherence** measures like C_v or NPMI evaluate whether the top words in each topic frequently co-occur in the corpus, which better captures interpretability.

**Human evaluation** remains the gold standard - do the discovered topics make sense to domain experts?

## Practical Workflow

Many practitioners run LDA with multiple K values (say, K ∈ {5, 10, 15, 20, 25}) and compare the results using both quantitative metrics and qualitative assessment of topic interpretability. The "right" K often depends on your specific use case and desired level of granularity.

TT: Here's another question.  Suppose I experimented with LDA by selecting news articles that I knew to fall into one of two primary topics: politics and business.  If I specified K=2 topics, how would LDA initiate the construction of the two clusters of documents?

Claude: Great experimental setup! LDA's initialization process is actually quite straightforward, though it has important implications for how the algorithm proceeds.

## Random Initialization

LDA typically starts by **randomly assigning each word token** in every document to one of the K topics. So with K=2, every word in every document gets randomly assigned to either "topic 1" or "topic 2" initially.

This means that at the start:

- Your political articles have words randomly split between both topics
- Your business articles also have words randomly split between both topics
- The initial topic-word distributions are essentially noise

## Why Random Assignment Works

Even though this seems chaotic, the Gibbs sampling process quickly corrects these random assignments. Here's the intuition:

**Political words cluster together:** Words like "election," "senator," "policy" appear frequently in your political documents. Even if initially scattered across both topics, the sampling process will tend to assign these co-occurring words to the same topic because it increases the likelihood.

**Business words cluster together:** Similarly, "revenue," "CEO," "market" co-occur in business documents, so they'll gravitate toward the same topic (likely the opposite one from politics words).

## The Self-Organization Process

Within just a few iterations, you'll typically see:

1. **Topic separation:** One topic starts accumulating political vocabulary, the other business vocabulary
2. **Document alignment:** Political articles develop higher probability for the "politics topic," business articles for the "business topic"
3. **Refinement:** The boundaries become sharper as the algorithm converges

## Alternative Initialization

Some implementations offer **informed initialization** where you can seed topics with specific words, but random initialization is standard because:

- It's simpler to implement
- It avoids bias from user assumptions
- The algorithm is robust enough to find good solutions from random starts

The key insight is that LDA's probabilistic framework naturally pushes co-occurring words toward the same topics, so even random initialization leads to meaningful clustering when clear thematic structure exists in your data.

TT: Thanks, that's very helpful. Now help me to understand the Gibbs sampler.

Claude: The Gibbs sampler is the computational engine that makes LDA inference tractable. It's a Markov Chain Monte Carlo (MCMC) method that samples from complex joint distributions by iteratively sampling from simpler conditional distributions.

## The Core Gibbs Sampling Principle

Instead of trying to sample from the complex joint posterior P(z, θ, φ | w, α, β) directly, Gibbs sampling breaks this into manageable pieces. It cycles through each variable, sampling from its conditional distribution given all other variables.

For LDA, the key insight is that we can **integrate out** the continuous parameters θ and φ analytically (thanks to Dirichlet-multinomial conjugacy), leaving us to sample only the discrete topic assignments z.

## The LDA Gibbs Update

For each word token w_{d,n} in document d at position n, we sample its topic assignment from:

P(z_{d,n} = k | z_{-d,n}, w, α, β) ∝ (n_{d,k}^{-d,n} + α_k) × (n_{k,w_{d,n}}^{-d,n} + β_{w_{d,n}}) / (n_k^{-d,n} + Σ_v β_v)

Where the superscript -d,n means "excluding the current token."

## Breaking Down This Formula

**First term (n_{d,k}^{-d,n} + α_k):** How often topic k appears in document d (plus prior). This captures document-topic affinity.

**Second term numerator (n_{k,w}^{-d,n} + β_w):** How often word w appears in topic k (plus prior). This captures topic-word affinity.

**Second term denominator (n_k^{-d,n} + Σ_v β_v):** Normalization ensuring topic k's word distribution sums to 1.

## Why This Works

The formula elegantly balances two forces:

- **Document consistency:** Prefer topics that are already prominent in this document
- **Topic coherence:** Prefer topics where this word commonly appears

## The Sampling Process

1. **Initialize:** Randomly assign each word token to a topic
2. **For each iteration:**
    - Update count matrices (n_{d,k}, n_{k,w}, n_k) based on current assignments
    - For each word token, sample new topic assignment using the formula above
    - Repeat until convergence (typically 1000-2000 iterations)

## Convergence Properties

The Markov chain has a unique stationary distribution equal to the posterior we want. After the burn-in period, samples approximate draws from P(z | w, α, β). We usually discard early samples and collect samples after convergence to estimate posterior quantities.

The beauty is that each local decision (reassigning one word) considers global statistics (document and topic patterns), allowing complex topic structures to emerge from simple local updates.

## References
