---
title: "Time Series Spectrum Analysis"
subtitle: "An Introduction"
author:
  - name: "Send comments to: Tony T (tthrall)"
date: last-modified
date-format: HH:mm ddd D-MMM-YYYY
editor: source
toc: true
toc-depth: 3
format:
  html:
    code: true
params:
  file_id: "s5c"
  file_prefix: "s5c_ts_fourier"
abstract:
  "Introduce the Fourier spectrum of a time series.  <br><br>"
bibliography: s5c_references.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo    = FALSE,
  error   = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
#| label: cran_libraries

library(assertthat)
library(astsa)
library(fable)
library(here)
library(imputeTS)
library(knitr)
library(latex2exp)
library(tidyverse)
library(tinytex)
library(tsibble)
library(tsibbledata)
library(xts)
library(zip)
```

```{r}
#| label: local_libraries

# library(ugrid)
```

```{r}
#| label: local_source

# city_temperatures (now moved to separate project)
# summaries and examples of daily temps from 300+ cities
# source(here("code", "city_temperatures.R"))
```

```{r}
#| label: HNL_NYC_t_imp

# daily temperatures at Honolulu and NYC, 1995 - 2020

# missing temperatures previously imputed as 
# variable t_imp via function imputeTS::na.ma()

HNL_NYC_t_imp <- readr::read_tsv(here::here(
  "data", "retain", "HNL_NYC_t_imp.txt"
)) |> 
  dplyr::mutate(across(
    .cols = c(Month, Day, Year), 
    .fns  = as.integer
  ))
```

```{r}
#| label: HNL_NYC_wide

# pivot wider for display
HNL_NYC_wide <- HNL_NYC_t_imp |> 
  dplyr::select(City, dt, temp) |> 
  tidyr::pivot_wider(
    id_cols = dt,
    names_from  = City, 
    values_from = temp
  ) |> 
  dplyr::rename(
    date = dt, 
    HNL  = Honolulu, 
    NYC  = `New York City`
  )
```

```{r}
#| label: HNL_NYC_smy

# stats for temps at Honolulu and NYC, 1995 - 2020

HNL_NYC_smy <- HNL_NYC_t_imp |>
    group_by(City) |>
    summarize(
      n_days = n(),
      n_miss = is.na(temp) |> sum() |> as.integer(),
      dt_min = min(dt, na.rm = TRUE),
      dt_max = max(dt, na.rm = TRUE),
      t_avg  = mean(temp, na.rm = TRUE),
      t_sd   = sd(temp, na.rm = TRUE)
    ) |>
    dplyr::ungroup()

# HNL_NYC_smy
# # A tibble: 2 × 7
#   City          n_days n_miss dt_min     dt_max     t_avg  t_sd
#   <chr>          <int>  <int> <date>     <date>     <dbl> <dbl>
# 1 Honolulu        9265     18 1995-01-01 2020-05-13  77.4  3.40
# 2 New York City   9265     20 1995-01-01 2020-05-13  56.0 17.1

# isolate stats for use in narrative
min_dt_HNL <- (HNL_NYC_smy$ dt_min) [[1]]
max_dt_HNL <- (HNL_NYC_smy$ dt_max) [[1]]
n_days_HNL <- (HNL_NYC_smy$ n_days) [[1]]

n_miss_HNL <- (HNL_NYC_smy$ n_miss) [[1]]
n_miss_NYC <- (HNL_NYC_smy$ n_miss) [[2]]

avg_HNL <- (HNL_NYC_smy$ t_avg) [[1]] |> round()
avg_NYC <- (HNL_NYC_smy$ t_avg) [[2]] |> round()

sd_HNL <- (HNL_NYC_smy$ t_sd) [[1]] |> round(digits = 1)
sd_NYC <- (HNL_NYC_smy$ t_sd) [[2]] |> round(digits = 1)

HNL_NYC_title <- "HNL-NYC temperatures $(^{\\circ} F)$"

```

## Introduction

The analysis of time series has proved to have both pragmatic and scientific value.  For time series that track business activity or broader economic activity, the ability to forecast future values with some degree of reliability has evident practical value.  The potential of such forecasting has influenced the development of time domain models, notably auto-regressive moving average (ARMA) models.

On the other hand, natural processes often exhibit periodic behavior, and have given rise to the Fourier decomposition of such time series, that is, the representation of the process as a sum of sine waves over a range of frequencies.

### Sunspots

An early example of such periodic behavior is shown in @fig-sunspots, the time series of monthly sunspot numbers.  Sunspots have been observed for millennia, but systematic scientific observations began in the 1600's.[^sunspot-history]

[^sunspot-history]: See [The Historical Sunspot Record](https://chandra.harvard.edu/edu/formal/icecore/The_Historical_Sunspot_Record.pdf).

Sunspots are a key indicator of the solar magnetic activity cycle in which the sun's magnetic polarity reverses every 11 years or so.  At the peak of a cycle the sun emits bursts of charged particles (solar flares) across the solar system that affect the Earth.  Consequently, understanding the solar cycle has both scientific and practical value.

Gaps between peak sunspot numbers vary, so a simple sinusoidal wave (or other function) having a fixed period does not fit the data well.  On the other hand, the variability of the data is clarified by Fourier decomposition and subsequent grouping of fitted sine waves within successive bands of frequencies. This approach is discussed in @sec-math-fmwk.

### Average Daily Temperatures

@fig-hnl-nyc-temps shows average daily temperatures for Honolulu and New York City over a 25-year period (1995-2020).  The overall average temperature in NYC is about `r avg_NYC` (F), well below that of Honolulu (`r avg_HNL`).  Also note that the standard deviation of NYC temperatures, roughly `r sd_NYC` (F), is well above that of Honolulu (`r sd_HNL`).  So Honolulu temperatures are higher than those of NYC and fluctuate less (as visitors to those two cities may testify).  Less clear, however, is the degree to which seasonal variation accounts for the overall variation in each city.  This is a question that can be addressed by Fourier decomposition.

## Mathematical Framework {#sec-math-fmwk}

### Data derived from a random process

Statistical applications are often based on one or more data frames in which each row represents an observation and each column represents a variable of interest.  For example, here are the first few values of the temperature data shown in @fig-hnl-nyc-temps.

```{r}
#| label: tbl-HNL-NYC-wide
#| tbl-cap: "Temperatures (F) from Honolulu and New York City, 1995-2020"

HNL_NYC_wide |> print(n = 3)
```

The distinction of time series analysis is that observations are indexed by time[^1001], and are not assumed to be statistically independent. Instead, time series data are typically modeled as a realization of a random process of the following form.[^real-valued]

[^1001]: Occasionally "time series" methods are applied to data indexed not by time but rather by length, or some other linear dimension.  For example see @thomson1977.

[^real-valued]: We assume the components of $X_\bullet(t)$ to be real-valued, unless stated otherwise.

$$
\begin{align}
  X_\bullet (t) = (X_1 (t), \ldots, X_d (t)) \in \mathbb{R}^d
\end{align}
$$ {#eq-X-bullet}

If the number of data columns is two or greater $(d \ge 2)$, the process is said to be multivariate. Otherwise, if $d = 1$ the process is said to be univariate, and the notation is simplified to $X (t)$.[^uni-subscripts]

[^uni-subscripts]: For a univariate time series $X(t)$ we denote distributional parameters either omitting a subscript or else using the subscript $X$.

In our example there are two data columns (`HNL`, `NYC`), so that $d = 2$, the unit of time is one day, and the sampling frequency is once per day.

The random process is idealized to span all time $(t \in \mathbb{Z})$, but the data of course span some finite period, $\tau_T$, of length $T$.

$$
\begin{align}
  \tau_T &=  \{ t_0, \; t_0 + 1, \ldots, t_f \} \\
  t_0 &=  \text{ initial data index} = \min \tau_T \\ 
  t_f &=  \text{ final data index} = \max \tau_T \\ 
  T &=  \text{ number of observatons } = t_f - t_0 + 1 \\ 
  \nu_T &= \tau_T - t_0 =  \{ 0, 1, \ldots, T-1 \}
\end{align}
$$ {#eq-tau-T}

The expected value of the random process may be modeled by various functions of time: a constant, a linear trend, a seasonal component (periodic function), etc.

$$
\begin{align}
  m_\bullet (t) &= E \{ X_\bullet (t) \}
\end{align}
$$ {#eq-m-bullet}

### Stationarity

In an additive model (which is most common), the residual random process, $X_\bullet (t) - m_\bullet (t)$ , is assumed to be **second-order stationary**.[^stationary] That is, if we shift $X_\bullet (\cdot)$ by any number of time units $s$ to obtain a new process $Y_\bullet (\cdot)$, we assume that the respective covariance structures of $X_\bullet (\cdot)$ and $Y_\bullet (\cdot)$ are the same.

[^stationary]: The assumption of second-order (or wide-sense) stationarity suffices for most time series models used in practice, but some cases may call for the assumption of *strict stationarity*. This means that for any finite set of times $(t_1, \ldots, t_K)$ and any time-shift $s$, the joint probability distribution of $(X_\bullet (t_1), \ldots, X_\bullet (t_K))$ is identical to that of $(X_\bullet (t_1 - s), \ldots, X_\bullet (t_K - s))$.

$$
\begin{align}
  Y_\bullet (t) &= \mathcal{B}^s \{ X_\bullet (\cdot) \} (t)  \\ 
  &= X_\bullet (t - s)
\end{align}
$$ {#eq-back-shift-s}

Here $\mathcal{B}$ denotes the *back-shift* operator that shifts time by one unit, so that $\mathcal{B}^s$ (that is, $s$ repeated applications of $\mathcal{B}$) shifts time by $s$ units. Then second-order stationarity can be expressed as follows.

$$
\begin{align}
  Cov \{ X_a (t + u), X_b (t) \} 
  &= Cov \{ Y_a (t + u), Y_b (t) \} \\ 
  &= Cov \{ X_a (t + u - s), X_b (t - s) \} \\ 
  \\ 
  & \text{for all } s \in \mathbb{Z} \text{ and } a, b \in \{ 1, \ldots, d \}
\end{align}
$$

### ACF: autocorrelation function

Setting $s = t$ in the last expression we have

$$
\begin{align}
  Cov \{ X_a (t + u), X_b (t) \} 
  &= Cov \{ X_a (u), X_b (0) \} 
\end{align}
$$ {#eq-stationarity-2}

Consequently, second-order stationarity enables one to define and estimate the following *auto-covariance* function.

$$
\begin{align}
  \gamma_{\bullet, \bullet} (u) &= \left \{ \gamma_{a, b} (u) \right \}_{a, b = 1}^d \\ 
  \\ 
  \text{where} \\
  \gamma_{a, b} (u) &= Cov \{ X_a (t + u), X_b (t) \} \\
  &= Cov \{ X_a (u), X_b (0) \} 
\end{align}
$$ {#eq-ACovF}

Although $\gamma_{\bullet, \bullet} (u)$ is a symmetric matrix at $u = 0$, it is not generally symmetric at $u \ne 0$. The symmetry relation we do have is that matrix $\gamma_{\bullet, \bullet} (-u)$ is the transpose of matrix $\gamma_{\bullet, \bullet} (u)$.

$$
\begin{align}
  \gamma_{a, b} (-u) &= Cov \{ X_a (t - u), X_b (t) \} \\
  &= Cov \{ X_a (t), X_b (t + u) \}  \\
  &= Cov \{ X_b (t + u), X_a (t) \}   \\
  &= \gamma_{b, a} (u)
\end{align}
$$ {#eq-ACovF-symm}

In particular, $\gamma_{a, a} (\cdot)$ is an even function: $\gamma_{a, a} (-u) = \gamma_{a, a} (u)$.

Consequently, the time-reversed process $Y_\bullet (t) = X_\bullet (-t)$ has the same auto-covariance structure as the original process, allowing for this matrix transposition.

The *autocorrelation (ACF)* function $\rho_{\bullet, \bullet}(\cdot)$ is the following scaled version of the auto-covariance function.

$$
\begin{align}
\rho_{\bullet, \bullet} (u) &= \left \{ \rho_{a, b} (u) \right \}_{a, b = 1}^d \\ 
  \\ 
  \text{where} \\
  \rho_{a, b} (u) &= corr \{ X_a (u), X_b (0) \} \\ 
  &= \frac{\gamma_{a, b} (u)}{\sqrt{\gamma_{a, a} (0) \; \gamma_{b, b} (0)}}
\end{align}
$$ {#eq-ACF}

### Spectrum

The term "spectrum" in its modern sense was introduced by Sir Isaac Newton[^art-Newton-spectrum], who used a prism to decompose sunlight into a rainbow of colors (light of different wavelengths or frequencies).  The term came to mean the intensity[^intensity-power] (denoted "I") of light (or other wave phenomena) as a function of wavelength or frequency.  This note uses "spectrum" in this sense (although the term has been further generalized).

[^art-Newton-spectrum]: See @Newton1672spectrum.

[^intensity-power]: In other contexts the term "intensity" is replaced by "power".  @wiener1930GHA introduced the term "power spectrum" of a signal.

### Spectral density functions

The _spectral density function_, $f_{\bullet, \bullet} (\lambda)$, of the random process $X_\bullet(\cdot)$ is defined as the Fourier transform of the auto-covariance function $\gamma_{\bullet, \bullet} (u)$.

$$
\begin{align}
  f_{\bullet, \bullet} (\lambda) &= \sum_{u = -\infty}^\infty \gamma_{\bullet, \bullet} (u) \times e^{- i \lambda u}
\end{align}
$$ {#eq-spec-dens}

where we assume that the elements of $\gamma_{\bullet, \bullet} (\cdot)$ are absolutely summable.

$$
\begin{align}
  \sum_{u = -\infty}^\infty \left | \; \gamma_{a, b} (u) \; \right | &< \infty \\ 
  &\text{for } a, b \in \{1, \ldots, d \}
\end{align}
$$ {#eq-ACF-sum}

@eq-ACF-sum is an example of a "mixing" condition.  It implies that $X_\bullet(t)$ and $X_\bullet(t + u)$ have vanishingly small correlation as $|u| \rightarrow \infty$.

Setting $b = a$, we obtain $f_{a, a} (\lambda)$, called the _power spectrum_ of $X_a (\cdot)$.  If $b \ne a$ then $f_{a, b} (\lambda)$ is called the _cross-spectrum_ of $\{ X_a (\cdot), X_b (\cdot) \}$.

From @eq-spec-dens we see that $f$ is periodic with period $2 \pi$.

$$
\begin{align}
  f_{\bullet, \bullet} (\lambda + 2 \pi) &= f_{\bullet, \bullet} (\lambda)
\end{align}
$$ {#eq-spec-periodic}

Due to the symmetries of the auto-covariance function (@eq-ACovF-symm), the spectral density matrix is Hermitian, that is, equal to its own conjugate transpose.

$$
\begin{align}
  f_{\bullet, \bullet} (\lambda)^* &= \sum_{u = -\infty}^\infty \gamma_{\bullet, \bullet} (u)^* \times e^{i \lambda u} \\ 
  &= \sum_{u = -\infty}^\infty \gamma_{\bullet, \bullet} (-u) \times e^{i \lambda u} \\ 
  &= \sum_{\nu = -\infty}^\infty \gamma_{\bullet, \bullet} (\nu) \times e^{- i \lambda \nu} \\ 
  &= f_{\bullet, \bullet} (\lambda)
\end{align}
$$ {#eq-spec-hermitian}

We also have that matrix $f_{\bullet, \bullet} (- \lambda)$ is the transpose of $f_{\bullet, \bullet} (\lambda)$.

$$
\begin{align}
  f_{\bullet, \bullet} (- \lambda) &= \sum_{u = -\infty}^\infty \gamma_{\bullet, \bullet} (u) \times e^{i \lambda u} \\ 
  &= \sum_{u = -\infty}^\infty \gamma_{\bullet, \bullet} (-u)^\prime \times e^{i \lambda u} \\ 
  &= \sum_{\nu = -\infty}^\infty \gamma_{\bullet, \bullet} (\nu)^\prime \times e^{- i \lambda \nu} \\ 
  &= f_{\bullet, \bullet} (\lambda)^\prime
\end{align}
$$ {#eq-spec-transpose}

That is 

$$
\begin{align}
  f_{a, b} (- \lambda) &= f_{b, a} (\lambda)
\end{align}
$$ {#eq-spec-elt-transpose}

Setting $b = a$ we see that the power spectrum is an even function.

$$
\begin{align}
  f_{a, a} (- \lambda) &= f_{a, a} (\lambda)
\end{align}
$$ {#eq-poweer-spec-even}

Lastly, @eq-spec-dens has the following inverse.

$$
\begin{align}
  \gamma_{\bullet, \bullet} (u) &= \int_{- \pi}^\pi f_{\bullet, \bullet} (\lambda) \times e^{i \lambda u} \; d\lambda
\end{align}
$$ {#eq-ACovF-from-spec-dens}

Setting $u = 0$ we see that the variance-covariance matrix $\gamma_{\bullet, \bullet} (0)$ of $X_\bullet (t)$ is the integral of $f_{\bullet, \bullet} (\cdot)$.

$$
\begin{align}
  \gamma_{\bullet, \bullet} (0) &= \int_{- \pi}^\pi f_{\bullet, \bullet} (\lambda) \; d\lambda
\end{align}
$$ {#eq-cov-from-spec-dens}

### Cramér representation

It turns out[^Cramér-rep] that @eq-ACF-sum implies the following Cramér representation of a centered version of $X_\bullet (\cdot)$.

[^Cramér-rep]: See @brillinger2001 [sec. 4.6].

$$
\begin{align}
  X_\bullet (t) - m_\bullet (t) &= \int_{0}^{2 \pi} e^{i \lambda t} \; dZ_\bullet (\lambda)
\end{align}
$$ {#eq-Cramér-rep}

where $Z_\bullet (\cdot)$ is a random complex-valued measure having mean value zero and having orthogonal increments[^conj-transp]:

[^conj-transp]: Given a complex-valued matrix or vector $\zeta$ we denote by $\zeta^*$ its conjugate transpose.

$$
\begin{align}
  E \left \{ Z_\bullet ( \Delta_1 ) \; Z_\bullet ( \Delta_2 )^* \right \} &= 0 \\ \\ 
  \text{whenever } \\ \\ 
  \Delta_1 \; \cap \; \Delta_2 &= \emptyset
\end{align}
$$ {#eq-ortho-incr}

The magnitude of $dZ_\bullet (\lambda)$ (which has mean zero) can be measured by the following complex-valued version of variance.

$$
\begin{align}
  E \left \{ dZ_\bullet ( \lambda ) \; dZ_\bullet ( \lambda )^* \right \} &= f_{\bullet, \bullet} (\lambda) \; d \lambda 
\end{align}
$$ {#eq-var-Z}

### Filtering

Suppose that $X(t)$ is a univariate random process satisfying @eq-ACF-sum and having mean zero.  Let $a(\nu)$ be a non-random summable sequence, and let $Y(\cdot)$ denote the convolution of $a(\cdot)$ with $X(\cdot)$.

$$
\begin{align}
  Y(\cdot) &= a(\cdot) * X(\cdot) \\ 
  \\ 
  &\text{so that } \\ 
  \\ 
  Y(t) &= \sum_{\nu = - \infty}^\infty a(\nu) \times X(t - \nu)
\end{align}
$$ {#eq-filter-demo}

Then $Y(\cdot)$ is said to be a _filtered_ version of $X(\cdot)$.  One can show[^filtered-Cramér] that $Y(\cdot)$ has the following Cramér representation.

[^filtered-Cramér]: See @brillinger2001 [sec. 4.6]

$$
\begin{align}
  Y (t) &= \int_{0}^{2 \pi} e^{i \lambda t} \; dZ_Y (\lambda) \\ 
  &= \int_{0}^{2 \pi} e^{i \lambda t} \; A(\lambda) \; dZ_X (\lambda)
\end{align}
$$ {#eq-filtered-Cramér}

where $A(\lambda)$ is the Fourier transform of $a(\nu)$.

$$
\begin{align}
  A(\lambda) &= \sum_{\nu = - \infty}^\infty a (\nu) \times e^{- i \lambda \nu}
\end{align}
$$ {#eq-transf-fn}

The Fourier transform $A (\lambda)$ is called the _transfer function_ of the filter, and the function $a(\nu)$ is called the _impulse response_ function of the filter.  (If one replaces $X (\cdot)$ with an "impulse" equal to unity at zero and equal to zero elsewhere one obtains $Y(t) = a(t)$.)

From @eq-var-Z we see that 

$$
\begin{align}
  f_{Y, Y} (\lambda) \; d \lambda &= E \left \{ dZ_Y ( \lambda ) \; dZ_Y ( \lambda )^* \right \}  \\ 
  &= E \left \{ A(\lambda) \; dZ_X ( \lambda ) \; dZ_X ( \lambda )^* \; A(\lambda)^* \right \}  \\ 
  &= \left | A(\lambda) \right |^2 \; E \left \{ dZ_X ( \lambda ) \; dZ_X ( \lambda )^* \right \}  \\ 
  &= \left | A(\lambda) \right |^2 \; f_{X, X} (\lambda) \; d \lambda \\ \\ 
  \text{so that } \\ \\ 
  f_{Y, Y} (\lambda) &= \left | A(\lambda) \right |^2 \; f_{X, X} (\lambda)
\end{align}
$$ {#eq-var-Z-Y}

### White noise

Let $W(\cdot)$ denote a univariate "white noise" process, that is, a process having zero mean and having the following auto-covariance function.

$$
\begin{align}
  \gamma_W (u) &= 
  \begin{cases}
    \sigma_W^2 & \text{for } u = 0 \\ 
    0 & \text{for } u \ne 0
  \end{cases}
\end{align}
$$ {#eq-ACF-wn}

We use the following notation to specify a white noise process.

$$
\begin{align}
  W &\sim wn(0, \sigma_W^2)
\end{align}
$$ {#eq-wn-notation}

The power spectrum density function, $f_{W, W} (\lambda)$ is readily seen to evaluate to the constant $\sigma_W^2$.

$$
\begin{align}
  f_{W, W} (\lambda) &= \sum_{u = -\infty}^\infty \gamma_W (u) \times e^{- i \lambda u} \\ 
  &= \gamma_W (0) \\ 
  &= \sigma_W^2
\end{align}
$$ {#eq-wn-spec-dens}

### White noise filtered

Now suppose that $X(\cdot)$ can be represented as the following filtered version of $W(\cdot)$.

$$
\begin{align}
  X(t) &= \sum_{\nu = 0}^\infty \psi_\nu \times W(t - \nu) \\ \\ 
  \text{where } \\ \\ 
  \sum_{\nu = 0}^\infty | \; \psi_\nu \; | &< \infty
\end{align}
$$ {#eq-causal-form}

From @eq-var-Z-Y we see that 

$$
\begin{align}
  f_{X, X} (\lambda) &= \left | \psi(e^{-i \lambda}) \right |^2 \; \sigma_W^2
\end{align}
$$ {#eq-spec-dens-causal}

where $\psi(\cdot)$ denotes the power series having coefficients $\psi (\nu)$.

$$
\begin{align}
  \psi (z) &= \sum_{\nu = 0}^\infty \psi_\nu \times z^\nu \quad \text{for } |z| \le 1
\end{align}
$$ {#eq-psi-power-series}

### ARMA models

To elaborate, let's define auto-regressive moving average (ARMA) models to develop such filtered versions of the white noise process $W(\cdot)$.  These models are defined by polynomials $\phi(\cdot)$ and $\theta(\cdot)$ applied to the back-shift operator $\mathcal{B}$ as follows.  (We set the mean of each process to zero in order to simplify the notation.)

$$
\begin{align}
  \phi(\mathcal{B}) \; X(\cdot) &= \theta(\mathcal{B}) \; W(\cdot) \\ \\
  \text{with } \\ \\ 
  \phi(z) &= 1 - \sum_{j = 1}^p \phi_j \; z^j \\ 
  \theta(z) &= 1 + \sum_{k = 1}^q \theta_k \; z^k \\ \\
  \text{so that } \\ \\ 
  X(t) &= \sum_{j = 1}^p \phi_j \; X(t - j) \; + \; W(t) \; + \; \sum_{k = 1}^q \theta_k \; W(t - k)
\end{align}
$$ {#eq-ARMA-p-q}

Taking the multiplicative inverse of $\phi(\cdot)$ we can re-express this model as the following filtering of the white noise process $W(\cdot)$.

$$
\begin{align}
  X(\cdot) &= \phi(\mathcal{B})^{-1} \;  \theta(\mathcal{B}) \; W(\cdot) \\ \\
  &= \psi(\mathcal{B}) \; W(\cdot) \\ \\ 
  \text{where } \\ \\ 
  \psi (z) &= \frac{\theta(z)}{\phi(z)} = \sum_{\nu = 0}^\infty \psi_\nu \times z^\nu 
\end{align}
$$ {#eq-ARMA-p-q-causal}

The power-series expansion of the rational function $\psi(\cdot)$ requires the roots of polynomial $\phi(\cdot)$ to all have magnitude greater than unity, which we assume to be the case.  Now let us specify some simple examples.

### AR(1)

First we define $X(\cdot)$ to be the following simple auto-regressive model, abbreviated $AR(1)$.

$$
\begin{align}
  X(t) &= \phi_1 \; X(t - 1) \; + \; W(t) \quad \text{with } | \phi_1 | < 1
\end{align}
$$ {#eq-AR-1}

so that 

$$
\begin{align}
  \psi (z) &= \frac{1}{1 - \phi_1 z} \\ 
  &= \sum_{\nu = 0}^\infty \phi_1^\nu \times z^\nu
\end{align}
$$ {#eq-AR-1-psi}

and 

$$
\begin{align}
  f_{X, X} (\lambda) &= \left | \psi(e^{-i \lambda}) \right |^2 \; \sigma_W^2 \\ 
  &= \left | \frac{1}{1 - \phi_1 e^{-i \lambda}} \right |^2 \; \sigma_W^2 \\ 
  &= \frac{\sigma_W^2}{1 - 2 \phi_1 \cos(\lambda) + \phi_1^2}
\end{align}
$$ {#eq-AR-1-spec}

### MA(1)

Next, let $Y(\cdot)$ denote the following simple moving average model $(MA(1))$.

$$
\begin{align}
  Y(t) &= W(t) \; + \; \theta_1 W(t - 1) \quad \text{with } | \theta_1 | < 1
\end{align}
$$ {#eq-MA-1}

so that 

$$
\begin{align}
  \psi (z) &= 1 + \theta_1 z
\end{align}
$$ {#eq-MA-1-psi}

and 

$$
\begin{align}
  f_{Y, Y} (\lambda) &= \left | \psi(e^{-i \lambda}) \right |^2 \; \sigma_W^2 \\ 
  &= \left | 1 + \theta_1 e^{-i \lambda} \right |^2 \; \sigma_W^2 \\ 
  &= \left ( 1 + 2 \theta_1 \cos(\lambda) + \theta_1^2 \right ) \; \sigma_W^2 \\ 
\end{align}
$$ {#eq-MA-1-spec}

### ARMA(1, 1)

Finally we let $Z(\cdot)$ denote the following $(ARMA(1, 1))$ model.

$$
\begin{align}
  Z(t) &= \phi_1 Z(t - 1) \; + \; W(t) \; + \; \theta_1 W(t - 1)
\end{align}
$$ {#eq-ARMA-1-1}

so that 

$$
\begin{align}
  \psi (z) &= \frac{1 + \theta_1 z}{1 - \phi_1 z} \\ 
  &= (1 + \theta_1 z) \; \sum_{\nu = 0}^\infty \phi_1^\nu \times z^\nu \\ 
  &= 1 \; + \; \sum_{k = 1}^\infty (\phi_1^k + \theta_1 \; \phi_1^{k - 1}) \times z^k
\end{align}
$$ {#eq-ARMA-1-1-psi}

and 

$$
\begin{align}
  f_{Z, Z} (\lambda) &= \left | \psi(e^{-i \lambda}) \right |^2 \; \sigma_W^2 \\ 
  &= \left | \frac{1 + \theta_1 e^{-i \lambda}}{1 - \phi_1 e^{-i \lambda}} \right |^2 \; \sigma_W^2 \\ 
  &= \frac{1 + 2 \theta_1 \cos(\lambda) + \theta_1^2}{1 - 2 \phi_1 \cos(\lambda) + \phi_1^2} \; \sigma_W^2
\end{align}
$$ {#eq-ARMA-1-1-spec}

### Comparison of ARMA models

```{r}
#| label: ARMA-sim-params

seed_sim <- (2L)^(7L) - 1L # 127, a Mersenne prime

phi_sim   <- 0.9
theta_sim <- 0.9

wn_sd <- 1

n_sim <- (2L)^(10L)
```

```{r}
#| label: ARMA-sim

set.seed(seed_sim)

sim_tbl <- tibble::tibble(
  idx = 1:n_sim, 
  W   = astsa::sarima.sim(n = n_sim, burnin = 0L), 
  X   = astsa::sarima.sim(n = n_sim, burnin = 0L, 
                          ar = phi_sim, innov = W), 
  Y   = astsa::sarima.sim(n = n_sim, burnin = 0L, 
                          ma = theta_sim, innov = W), 
  Z   = astsa::sarima.sim(n = n_sim, burnin = 0L, 
                          ar = phi_sim, ma = theta_sim, innov = W)
)
```

```{r}
#| label: ARMA-sim-long

sim_long <- sim_tbl |> 
  tidyr::pivot_longer(
    cols = W:Z, 
    names_to = "var"
  ) |> 
  dplyr::mutate(
    tag = case_when(
      var == "W" ~ "wn", 
      var == "X" ~ "AR_1", 
      var == "Y" ~ "MA_1", 
      var == "Z" ~ "ARMA_1_1")
  ) |> 
  dplyr::select(idx, var, tag, value)
```

@fig-ARMA-sim-ts shows the respective time series simulated from these simple ARMA models.


```{r}
#| label: fig-ARMA-sim-ts
#| fig-cap: "Simulation from simple ARMA models"

g_sim_long <- sim_long |> 
  dplyr::group_by(var) |> 
  ggplot2::ggplot(mapping = aes(
    x = idx, y = value, color = var, fill = var
  )) + 
  ggplot2::geom_line(show.legend = FALSE) + 
  ggplot2::facet_grid(
    rows   = vars(var), 
    scales = "free_y"
  ) + 
  ggplot2::labs(
    title = "Simulated ARMA models", 
    subtitle = "W = White Noise, X = AR(1), Y = MA(1), Z = ARMA(1, 1)"
  ) + 
  ggplot2::xlab("time index")
g_sim_long
```

The trajectory of the white noise process $W(\cdot)$ is indeed noisy; it would remain qualitatively unchanged by any permutation of its values.  The MA(1) model generating $Y(\cdot)$ also displays a rather noisy trajectory.  The remaining models, the AR(1) model generating $X(\cdot)$ and the ARMA(1, 1) model generating $Z(\cdot)$ show greater continuity.

@fig-ARMA-spec shows the respective spectral density functions of these ARMA models.  The spectral densities of the AR(1) model and the ARMA(1, 1) model are concentrated at lower frequencies, correponding to greater wavelengths.  Note that the horizontal axis labeled "freq" is a scaled version of $\lambda$, namely $\mathcal{f} = \frac{\lambda}{2 \pi}$, which is the inverse of wavelength.

```{r}
#| label: ARMA-model-spec

wn_spec_lst <- astsa::arma.spec(
    var.noise = wn_sd^2, 
    n.freq    = n_sim/2L, 
    plot      = FALSE)
freq_model <- wn_spec_lst$ freq

# extract spectral density from listed results
spec_arma_spec <- function(
    ar = 0, # AR coefficients of z^j, j > 0  
    ma = 0  # MA coefficients of z^k, k > 0  
) {
  spec_lst <- astsa::arma.spec(
    ar        = ar, 
    ma        = ma, 
    var.noise = wn_sd^2, 
    n.freq    = n_sim/2L, 
    plot      = FALSE)
  return(spec_lst$ spec)
}

arma_spec_tbl <- tibble::tibble(
  freq = freq_model, 
  
  W = spec_arma_spec(), 
  X = spec_arma_spec(ar = phi_sim), 
  Y = spec_arma_spec(ma = theta_sim), 
  Z = spec_arma_spec(
    ar = phi_sim, 
    ma = theta_sim)
)
```

```{r}
#| label: arma-spec-long

arma_spec_long <- arma_spec_tbl |> 
  tidyr::pivot_longer(
    cols = W:Z, 
    names_to = "var", 
    values_to = "spec"
  ) |> 
  dplyr::mutate(
    tag = case_when(
      var == "W" ~ "wn", 
      var == "X" ~ "AR_1", 
      var == "Y" ~ "MA_1", 
      var == "Z" ~ "ARMA_1_1")
  ) |> 
  dplyr::select(freq, var, tag, spec)
  
```

```{r}
#| label: fig-ARMA-spec
#| fig-cap: "Spectral density function of each ARMA model"

g_arma_spec <- arma_spec_long |> 
  dplyr::group_by(var) |> 
  ggplot2::ggplot(mapping = aes(
    x = freq, y = spec, color = var, fill = var
  )) + 
  ggplot2::geom_line(show.legend = FALSE) + 
  ggplot2::facet_grid(
    rows   = vars(var), 
    scales = "free_y"
  ) + 
  ggplot2::labs(
    title = "Spectral density of simple ARMA models", 
    subtitle = "W = White Noise, X = AR(1), Y = MA(1), Z = ARMA(1, 1)"
  ) + 
  ggplot2::ylab("spectral density")
g_arma_spec
```



```{r}
#| label: ARMA-sim-spec-est

spec_W_lst <- sim_tbl$ W |> 
  stats::spec.pgram(detrend = FALSE, plot = FALSE, taper = 1/4)
freq_sim   <- spec_W_lst$ freq
spec_W     <- spec_W_lst$ spec

# extract estimated spectral density from listed results
spec_pgram_spec <- function(
  x # univariate time series  
  ) {
  spec_lst <- x |> 
    stats::spec.pgram(
      detrend = FALSE, 
      plot    = FALSE, 
      taper   = 1/4, 
      spans   = c(31, 31))
  return(spec_lst$ spec)
}

spec_sim <- tibble::tibble(
  freq = freq_sim, 
  
  W = sim_tbl$ W |> spec_pgram_spec(), 
  X = sim_tbl$ X |> spec_pgram_spec(), 
  Y = sim_tbl$ Y |> spec_pgram_spec(), 
  Z = sim_tbl$ Z |> spec_pgram_spec()
)

```

```{r}
#| label: spec-sim-long

spec_long <- spec_sim |> 
  tidyr::pivot_longer(
    cols = W:Z, 
    names_to = "var", 
    values_to = "spec"
  ) |> 
  dplyr::mutate(
    tag = case_when(
      var == "W" ~ "wn", 
      var == "X" ~ "AR_1", 
      var == "Y" ~ "MA_1", 
      var == "Z" ~ "ARMA_1_1")
  ) |> 
  dplyr::select(freq, var, tag, spec)
  
```

```{r}
#| label: fig-ARMA-sim-spec-est
#| fig-cap: "Estimated spectral density function of each ARMA simulation"
#| eval: false

g_spec_sim <- spec_long |> 
  dplyr::group_by(var) |> 
  ggplot2::ggplot(mapping = aes(
    x = freq, y = spec, color = var, fill = var
  )) + 
  ggplot2::geom_line(show.legend = FALSE) + 
  ggplot2::facet_grid(
    rows   = vars(var), 
    scales = "free_y"
  ) + 
  ggplot2::labs(
    title = "Spectral density estimates of simulated ARMA models", 
    subtitle = "W = White Noise, X = AR(1), Y = MA(1), Z = ARMA(1, 1)"
  )
g_spec_sim
```

## Closing Remarks

## A: Glossary of Math Symbols

$$
\begin{align}
  {a_{\bullet, \bullet} (u)} 
    \quad & \quad \text{matrix of filter coefficients at time-shift } u \\ 
  \\ 
  \mathcal{B} 
     \quad & \quad \text{back-shift operator } \\ 
  \\ 
  \nabla 
     \quad & \quad \text{difference operator } \mathcal{I - B} \\ 
  \\ 
  \nabla_\sigma 
     \quad & \quad \text{seasonal difference operator } \mathcal{I - B}^\sigma \\ 
  \\ 
  \Delta_u
     \quad & \quad \{ 1, 2, \ldots, u-1 \} \text{ for } u \ge 2 \\ 
  \\ 
  \phi(\cdot)
     \quad & \quad \text{AR polynomial, } 1 - \sum_{\nu = 1}^p \phi_\nu \; z^\nu \\ 
  \\ 
  \Gamma 
     \quad & \quad \text{auto-covariance matrix, } \Gamma [j, \nu] = \gamma_X (\nu - j) \text{, for } j, \nu \in \nu_T \\ 
  \\ 
  \gamma_{\bullet, \bullet} (u) 
     \quad & \quad \text{auto-covariance matrix at time-shift } u \\ 
  \\ 
  \mathcal{I} 
     \quad & \quad \text{identity operator } \\ 
  \\ 
  MSE 
     \quad & \quad \text{mean squared error } \\ 
  \\ 
  m_\bullet (t) 
     \quad & \quad E \{ X_\bullet (t) \} \text{, expected value of } X_\bullet (t) \\ 
  \\ 
  \mu_X 
     \quad & \quad E\{ X (t_0) \} \text{, expected value of } X (t_0) \\ 
  \\   
  \hat{\mu}_X
     \quad & \quad \text{sample mean, } S_{X, T} / T \\ 
  \\   
  \mathcal{N}(\cdot)
     \quad & \quad \text{a univariate, stationary, Gaussian process having mean 0} \\ 
  \\   
  \tilde{\mathcal{N}} (u + t_f \; | \; \wp (t_f))
     \quad & \quad E \left \{ \mathcal{N} (u + t_f) \; | \; \mathcal{N} (s), \; s \le t_f \right \} \\ 
  \\ 
  \nu_T 
     \quad & \quad \{ 0, 1, \dots, T-1 \} \\ 
  \\ 
  \wp (t_f) 
     \quad & \quad \{ s \in \mathbb{Z} \; | \; s \le t_f \} \\ 
  \\ 
  \rho_{\bullet, \bullet} (u) 
     \quad & \quad \text{autocorrelation matrix at time-shift } u \\ 
  \\ 
  \rho (u | u) 
     \quad & \quad corr(X(t+u) - \hat{X}(t+u \; | \; - \Delta_u), \; X(t) - \hat{X}(t \; | \; \Delta_u)) \\ 
  \\ 
  S_{X, T} 
     \quad & \quad \text{the sample sum of T observations of } X (\cdot) \\ 
  \\ 
  \sigma_X^2 
     \quad & \quad Var \{ X (t_0) \} \text{, variance of } X (t_0) \\ 
  \\ 
  T 
     \quad & \quad \text{number of time series observations } \\ 
  \\ 
  t_0 
     \quad & \quad \text{initial time index of time series observations } \\ 
  \\ 
  t_f 
     \quad & \quad \text{final time index of time series observations } \\ 
  \\ 
  \tau_T 
     \quad & \quad t_0 + \{ 0, 1, \ldots, T-1  \} \\ 
  \\ 
  \theta(\cdot) 
     \quad & \quad \text{MA polynomial, } 1 + \sum_{\nu = 1}^q \theta_\nu \; z^\nu \\ 
  \\ 
  W(\cdot) 
     \quad & \quad \text{white noise process, } W(\cdot) \sim wn(0, \sigma_W^2)  \\ 
  \\ 
  X_\bullet (t) 
     \quad & \quad (X_1 (t), \ldots, X_d (t)) \text{, a multivariate random process } \\ 
  \\ 
  \hat{X}(t \; | \; \Delta_u) 
     \quad & \quad \text{approximates } X(t) \text{ as affine function of } \{ X(t + \nu) \}_{\nu \in \Delta_u} \\ 
  \\ 
  \hat{X} (u + t_f \; | \; \wp (t_f)) 
     \quad & \quad \text{predicts } X(u + t_f) \text{ as affine function of } \{ X(t) \}_{t \in \wp (t_f)} \\ 
  \\ 
  \breve{X} (t \; | \; \tau_T) 
     \quad & \quad \text{trucnated predictor of } X(t) \text{ based on } \{ X(t) \}_{t \in \tau_T}
\end{align}
$$

## B: Data Examples

### Sunspot Numbers: Monthly Averages, 1749 - Present

#### Description

@fig-sunspots below shows `sunspot.month`, monthly average sunspot numbers, in `R` package `datasets`. Source: Solar Influences Data Analysis Center (SIDC), Royal Observatory of Belgium.

```{r}
#| label: fig-sunspots
#| fig-cap: "Sunspot numbers: monthly average"

g_sunspots <- astsa::tsplot(
  x = datasets::sunspot.month, 
  ylab = "sunspot number", 
  col = 4)
```

#### Context: Questions and Objectives

As described in [Recalibration of the Sunspot-Number: Status Report \| Solar Physics](https://link.springer.com/article/10.1007/s11207-023-02136-3), the record of sunspot numbers links past and present solar behavior, and is the primary input for reconstructions of total solar irradiance (TSI) for years before 1978.

### Daily City Temperatures: 1995-2020

#### Description

@fig-hnl-nyc-temps below shows the daily average temperature in Honolulu and New York City from `r min_dt_HNL` to `r max_dt_HNL`.[^hnl-nyc-impute]

[^hnl-nyc-impute]: For convenience missing values have been imputed (estimated) with a weighted running average spanning 7 days.  This was done with `R` function `imputeTS::na.ma()`.  Of the `r n_days_HNL` days shown, there were `r n_miss_HNL` missing values for Honolulu and `r n_miss_NYC` missing values for NYC.

```{r}
#| label: fig-hnl-nyc-temps
#| fig-cap: "Daily Average Temperature (F), HNL and NYC, 1995-2020"

g_HNL_NYC_t_imp <- HNL_NYC_t_imp |> 
  dplyr::group_by(City) |> 
  ggplot2::ggplot(mapping = aes(
    x = dt, y = t_imp, color = City, fill = City
  )) + 
  ggplot2::geom_line(show.legend = FALSE) + 
  ggplot2::facet_grid(rows = dplyr::vars(City)) + 
  ggplot2::labs(
    title = "Daily Average Temperature for HNL, NYC, 1995-2020") + 
  ggplot2::xlab("Year") + 
  ggplot2::ylab("Temperature (F)")
g_HNL_NYC_t_imp
```


#### Context: Questions and Objectives

The temperatures shown are available from [Kaggle](https://www.kaggle.com/datasets/sudalairajkumar/daily-temperature-of-major-cities) (obtained from the University of Dayton) as part of a collection of temperatures at 157 cities in the US and 167 cities across the globe.  These surface temperatures are one of several variables that are used to model weather systems worldwide.

Also see the following R packages:

  - [GSOD](https://cran.r-project.org/web/packages/GSODR/readme/README.html), Global Surface Summary of the Day
  - [climate](https://cran.r-project.org/web/packages/climate/vignettes/getstarted.html)

## References
