---
title: "The Dirichlet Distribution"
subtitle: "Selected topics from Part 1 of Data Mining Intro"
author: 
  - name: "Send comments to: Tony T (adthral)"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M', usetz = TRUE)`"
output: 
  html_document:
    toc: true
    df_print: paged
    mathjax: default
  word_document:
    toc: true
    df_print: tibble
  pdf_document:
    toc: true
    df_print: tibble
abstract: 
  "The Dirichlet and Multinomial distributions are introduced in preparation for a discussion of Latent Dirichlet Allocation (LDA)."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE, 
  error   = FALSE, 
  message = FALSE, 
  warning = FALSE
)
```

```{r libraries}
library(assertthat)
library(GGally)
# library(gutenbergr)
library(here)
# library(ISLR2)
# library(janeaustenr)
library(LaplacesDemon)
library(latex2exp)
# library(learnr)
# library(quanteda)
# library(SAPP)
# library(tidytext)
library(tidyverse)
# library(timeSeriesDataSets)
library(tinytex)
# library(tm)
# library(tokenizers)
library(topicmodels)
# library(tsibble)
library(tufte)

```

```{r local_source}
# source(here("code", "rmse_per_grp.R"))
# source(here("code", "xtabs_to_jaccard.R"))
# source(here("code", "gen_circuit.R"))
```

------------------------------------------------------------------------

## Background

Latent Dirichlet Allocation (LDA) was proposed as a method of topic modeling in 2003 in a paper by Blei, Ng, and Jordan.  The method is briefly mentioned in Part 1 of the course.  Several course participants requested a more detailed description.  This note prepares for the requested response by introducing the Dirichlet and Multinomial distributions.

## The Multinomial Distribution

Consider a categorical (qualitative) random variable $X$ that randomly selects one of $K$ distinct categories $\{c_1, \ldots, c_K \}$.

$$
\begin{align}
  X &\in (c_1, \ldots, c_K) \\ 
  \\ 
  p_k &= P(X = c_k) > 0 \\ 
  \\ 
  \sum_{k = 1}^K p_k &= 1 \\ 
\end{align}
$$

Probability vector $p_{\bullet} = (p_1, \ldots, p_K)$ is thus the probability distribution of $X$ on $\{c_1, \ldots, c_K \}$.

If the categories are ordered, we might represent $X$ numerically as a random index $k \in \{1, \ldots, K \}$.  But for present purposes we suppose the categories are not ordered, and we represent $X$ as a vector of random indicator variables[^one-hot].

[^one-hot]: In machine learning this mapping of a categorical variable to an indicator vector is called "one-hot encoding".

$$
\begin{align}
  I_{\bullet} &= (I_1, \ldots, I_K) \\ 
  \\ 
  I_k &= 1 \text{, if } X = c_k \\ 
  I_k &= 0 \text{, if } X \ne c_k  \\ 
  \\ 
  I_{\bullet} &= \mathbb{e}_1 = (1, 0, \ldots, 0)  \text{, with probability } p_1\\ 
  I_{\bullet} &= \mathbb{e}_2 = (0, 1, \ldots, 0)  \text{, with probability } p_2\\ 
  \vdots \\ 
  I_{\bullet} &= \mathbb{e}_K = (0, 0, \ldots, 1)  \text{, with probability } p_K \\ 
\end{align}
$$

That is, the random indicator vector $I_{\bullet}$ selects one of the Euclidean basis vectors $\mathbb{e}_k \in \mathbb{R}^K$ with probability $p_k$. Therefore the expected value of $I_{\bullet}$ equals the vector of category probabilities $p_{\bullet} = (p_1, \ldots, p_K)$.

$$
\begin{align}
  E(I_k) &= P(I_k = 1) \\ 
  & = P(X = c_k) \\ 
  & = p_k \\
  \\ 
  E(I_{\bullet}) &= (E(I_1), \ldots, E(I_K)) \\ 
  &= (p_1, \ldots, p_K) \\ 
  &= p_{\bullet} \\ 
\end{align}
$$

Now suppose that $(X_1, \ldots, X_n)$ are independent random variables all having the same distribution as $X$.  Corresponding to $X_{\nu}$ we have an indicator vector that we'll denote as $I_{\bullet}^{(\nu)}$.  Let $S_{\bullet}$ denote the sum over the $n$ indicator vectors $\{ I_{\bullet}^{(\nu)} \}_{\nu}$.

$$
\begin{align}
  S_{\bullet} &= \sum_{\nu = 1}^n I_{\bullet}^{(\nu)} \text{, so that } \\ 
  \\ 
  S_k &= \sum_{\nu = 1}^n I_k^{(\nu)} \\ 
  &= \text{ number of } \nu \text{ such that } X_{\nu} = c_k
\end{align}
$$

For any given set of possible counts $s_{\bullet} = (s_1, \ldots, s_K)$, that is, of non-negative integers summing to $n$, and for a given probability vector $p_{\bullet}$, the probability that $S_{\bullet} = s_{\bullet}$ is as follows.

$$
\begin{align}
  P(S_{\bullet} = s_{\bullet} \; | \; p_{\bullet}) &= {n \choose s_{\bullet}} \prod_{k = 1}^K p_k^{s_k}
\end{align}
$$

where 

$$
\begin{align}
  {n \choose s_{\bullet}} &= \frac{n!}{s_1! s_2! \cdots s_K!}
\end{align}
$$

gives the number of assignments of $n$ objects to the $K$ categories such that each category $c_k$ receives the prescribed number $s_k$ of objects.

The probability distribution of $S_{\bullet}$ is called the _multinomial distribution_.

For $K = 2$, this simplifies to the binomial distribution, with parameters $(n, p)$, where $(p_1, p_2) = (p, \; 1 - p)$.  If $\nu$ denotes the observed number of succeses in $n$ Bernoulli trials, then $(s_1, s_2) = (\nu, \; n - \nu)$.

## Bayesian Inference

The Bayesian approach to statistical estimation provides a framework for representing the state of knowledge, or degree of uncertainty, about a model parameter before and after collecting relevant data.

As an example consider the binomial distribution mentioned above, where we are estimating the probability $p$ of success based on a sequence of $n$ independent Bernoulli trials.  Following the notation above, we use two dependent indicator random variables $(I_1, I_2)$ to represent success and failure respectively, with $I_2 = 1- I_1$, and with $(p_1, p_2) = (p, \; 1-p)$.

Prior to observing the sequence of Bernoulli trials, we might represent the state of information about $p$ as a uniform distribution, so that each possible value of $p \in [0, 1]$ is deemed equally likely.  Alternatively, if the Bernoulli sequence represents the outcomes of tossing a coin that is presumed fair, or nearly fair, we might represent that information as a probability distribution having a mode at the value $p = 1/2$.

More specifically, it turns out to be mathematically convenient to represent prior information about probability $p$ as a member of the beta family of probability distributions over the unit interval.  The uniform distribution is the special case of setting beta shape parameters to the values $(1, 1)$.  Alternatively setting the shape parameters to $(2, 2)$ gives a density function symmetric about the mode at $p = 1/2$.

```{r beta_xmpl}
# choices of beta shape parameters (a, b)
ab_tbl   <- tibble::tibble(a = 1:2, b = 1:2)
n_shapes <- nrow(ab_tbl)

# number of plotting points for beta density function
n_pts <- 200L

# grid points along x axis
x_grid <- ((1:n_pts) - 0.5)/n_pts

# long tibble to represent different shapes of beta distribution
beta_xmpl <- tibble::tibble(
  p     = rep(x_grid, n_shapes), 
  shape = rep(1:n_shapes, each = n_pts), 
  alpha = rep(ab_tbl$ a[1:n_shapes], each = n_pts), 
  beta  = rep(ab_tbl$ b[1:n_shapes], each = n_pts), 
  f     = dbeta(p, alpha, beta)
)
```

```{r g_beta_xmpl}
g_beta_xmpl <- beta_xmpl |> 
  dplyr::mutate(shape = as_factor(shape)) |> 
  ggplot(mapping = aes(
    x = p, y = f, color = shape
  )) + 
  geom_line(show.legend = FALSE) + 
  facet_grid(row = vars(shape)) + 
  labs(
    title = "Alternative prior densities of parameter p", 
    subtitle = "Beta shape parameters (1, 1) and (2, 2)"
  )
g_beta_xmpl
```

In general, a beta distribution having shape parameters $(\alpha, \beta)$ has the following density function.

$$
\begin{align}
  P(p) &= \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)} p^{\alpha - 1} (1-p)^{\beta -1 } \\ 
  \\ 
  & \text{with }  \alpha > 0 \text{, and } \beta > 0
\end{align}
$$

Suppose now that we adopt the distribution just mentioned, $\mathcal{Beta}(\alpha, \beta)$, as the prior distribution of success probability $p$, for some specified positive parameter values $(\alpha, \beta)$.  We then observe $s_1$ successes and $s_2$ failures from $n = s_1 + s_2$ independent Bernoulli trials.  Based on these observations we update the prior distribution to form the posterior distribution of $p$ as follows.

$$
\begin{align}
  P(p \; | \; S_{\bullet} = s_{\bullet}) &= \frac{P(p, \;  S_{\bullet}=s_{\bullet})}{P(S_{\bullet}=s_{\bullet})} \\ 
  &= \frac{P(S_{\bullet}=s_{\bullet} \; | \; p) \; P(p)}{P(S_{\bullet}=s_{\bullet})} \\ 
\end{align}
$$

Note that 

$$
\begin{align}
  P(S_{\bullet}=s_{\bullet} \; | \; p) \; P(p) &= {n \choose s_1} p^{s_1} (1 - p)^{n - s_1} \times \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)} p^{\alpha - 1} (1-p)^{\beta -1 } \\ 
  &= {n \choose s_1} \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)} p^{s_1 + \alpha - 1} (1 - p)^{n - s_1 + \beta - 1} \\
  &= \frac{(s_1 + s_2)!}{s_1! s_2!} \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)} p^{\alpha + s_1 - 1} (1 - p)^{\beta + s_2 - 1} \\ 
\end{align}
$$

so that 

$$
\begin{align}
  P(S_{\bullet} = s_{\bullet}) &= \int_0^1 P(S_{\bullet}=s_{\bullet} \; | \; p) \; P(p) \,dp \\ 
  &= \frac{(s_1 + s_2)!}{s_1! s_2!} \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)} \int_0^1 p^{\alpha + s_1 - 1} (1 - p)^{\beta + s_2 - 1} \,dp \\ 
  &= \frac{(s_1 + s_2)!}{s_1! s_2!} \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha) \Gamma(\beta)} \frac{\Gamma(\alpha + s_1)\Gamma(\beta + s_2)}{\Gamma(\alpha + \beta + s_1 + s_2)} \\ 
\end{align}
$$

Consequently, the ratio of the last two expressions gives 

$$
\begin{align}
  P(p \; | \; S_{\bullet} = s_{\bullet}) &= \frac{P(p, \;  S_{\bullet}=s_{\bullet})}{P(S_{\bullet}=s_{\bullet})} \\ 
  &= \frac{P(S_{\bullet}=s_{\bullet} \; | \; p) \; P(p)}{P(S_{\bullet}=s_{\bullet})} \\ 
  &= \frac{\Gamma(\alpha + \beta + s_1 + s_2)}{\Gamma(\alpha + s_1) \Gamma(\beta + s_2)} p^{\alpha + s_1 - 1} (1-p)^{\beta + s_2 -1}
\end{align}
$$

That is, the posterior distribution is $\mathcal{Beta}(\alpha + s_1, \beta + s_2)$.

This is the "mathematical convenience" previously alluded to: the prior and posterior probability distributions of $p$ belong to the same family of parametric probability distributions, namely the $\mathcal{Beta}$ family.  For this reason the $\mathcal{Beta}$ family is said to be _conjugate_ to the binomial family.

## The Dirichlet Probability Distribution

### Definition

The binomial distribution is a special case of the multinomial distribution in which the number of categories $K$ is equal to 2.  (In the discussion above we referred to the categories as success and failue, respectively.)  More generally, for a fixed integer $K \ge 2$, the family of distributions conjugate to the multinomial family of distributions over $K$ categories is the following _Dirichlet_ family.

$$
\begin{align}
  P(p_{\bullet}) &= \frac{\Gamma(\alpha_1+ \cdots + \alpha_K)}{\Gamma(\alpha_1) \times \cdots \times \Gamma(\alpha_K)} \prod_{k = 1}^K p_k^{\alpha_k - 1} \\ 
  \\ 
  & \text{with }  \alpha_k > 0 \text{ for } k \in \{1, \ldots, K \}
\end{align}
$$

This is the $\mathcal{Dirichlet}(\alpha_{\bullet})$ distribution over $p_{\bullet}$, where probability vector $p_{\bullet}$ ranges over the $K - 1$ dimesional simplex such that each component $p_k$ is non-negative and all the components $(p_1, \cdots, p_K)$ together sum to unity.

### Special Case: $\alpha_k = \frac{\alpha_{+}}{K}$

We will denote the sum of the components of $\alpha_{\bullet} = (\alpha_1, \ldots, \alpha_K)$ as $\alpha_{+}$.[^alpha_plus]

[^alpha_plus]: An alternative notation for the sum of $(\alpha_1, \ldots, \alpha_K)$ is $\alpha_0$.

$$
\begin{align}
  \alpha_{+} &= \sum_{k = 1}^K \alpha_k \\
\end{align}
$$

Consider the special case in which all the components of $\alpha_{\bullet}$ have the same value 

$$
\begin{align}
  \alpha_k &= \frac{\alpha_{+}}{K} \\
  & \text{for } k \in \{1, \dots, K \}
\end{align}
$$

The probability distribution is then symmetric in the components of $p_{\bullet}$, and $\alpha_{+}$ is referred to as the _concentration parameter_.  The uniform distribution over the domain of $p_{\bullet}$ (that is, over the $K - 1$ dimensional simplex) is obtained by setting $\alpha_{+} = K$.  Setting $\alpha_{+} > K$ concentrates the distribution around the centroid 

$$
\begin{align}
  p_{\textbf{ctr}} &= (\frac{1}{K}, \ldots, \frac{1}{K}) \\
\end{align}
$$

For example the previous figure shows two alternative concentrations of the symmetric beta density function of scalar parameter $p$, where $(p_1, p_2) = (p, \; 1 - p)$, and $p_{\textbf{ctr}} = (\frac{1}{2}, \frac{1}{2})$.

Alternatively, setting $\alpha_{+} < K$ concentrates the distribution away from the centroid toward the corners of the simplex.

### Posterior Distribution

Recall that $S_{\bullet}$ constructed above has a multinomial distribution if it can be represented as the sum of $n$ independent trials, each trial yielding one of the Euclidean basis vectors $\mathbb{e}_k$ with probability $p_k$.  Then $S_k$, the $k^{th}$ component of $S_{\bullet}$, counts the number of trials yielding $\mathbb{e}_k$.

Now suppose that $p_{\bullet}$ has prior distribution $\mathcal{Dirichlet}(\alpha_{\bullet})$, for some specification of $\alpha_{\bullet}$, and that the observed outcome of the $n$ trials is a given vector $s_{\bullet}$ of counts.  What is the posterior distribution of $p_{\bullet}$ given the observation $S_{\bullet} = s_{\bullet}$?  This turns out to be 

$$
\begin{align}
  \{ p_{\bullet} \; | \; s_{\bullet} \} &\sim \mathcal{Dirichlet}(\alpha_{\bullet} + s_{\bullet})
\end{align}
$$

That is, the Dirichlet family is conjugate to the multinomial family, just as the beta family is conjugate to the binomial family.

## Resources

[Dirichlet distribution - Wikipedia](https://en.wikipedia.org/wiki/Dirichlet_distribution)

