---
title: "Conditional Distributions"
subtitle: "Part 1, session 1b of Data Mining Intro"
author: 
  - name: "Send comments to: Tony T (adthral)"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M', usetz = TRUE)`"
output: 
  html_document:
    toc: true
    df_print: paged
    mathjax: default
  word_document:
    toc: true
    df_print: tibble
  pdf_document:
    toc: true
    df_print: tibble
abstract: 
  "Review concepts and techniques of exploratory data analysis."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE, 
  error   = FALSE, 
  message = FALSE, 
  warning = FALSE
)
```

```{r libraries}
library(assertthat)
library(here)
library(latex2exp)
library(tidyverse)
library(tinytex)
library(UsingR)

```

```{r local_source}
# source(here("code", "xxx.R"))
```

------------------------------------------------------------------------

## Heights of Fathers and Sons

The box-plot below can be regarded as the (sample) *conditional distribution* of sons' heights given the height (to the nearest odd inch) of each son's father.

```{r father_son_ht}
father_son_ht <- UsingR::father.son |> 
  as_tibble() |> 
  rename(father = fheight, son = sheight)
```

```{r f_mpt}
# f_ivl: successive 2-inch intervals of father heights
# f_mpt: mid-point of each interval
father_son_ht <- father_son_ht |> 
  mutate(
    f_ivl = cut(father, seq(58, 76, 2)), 
    f_mpt = (2 * ceiling(father/2)) - 1
  )
```

```{r s_stats_per_f_mpt}
# son statistics per f_ivl
s_stats_per_f_mpt <- father_son_ht |>    
  group_by(f_ivl, f_mpt) |>    
  summarise(     
    s_count = n(),      
    s_min   = min(son, na.rm = TRUE),      
    s_max   = max(son, na.rm = TRUE),      
    s_avg   = mean(son, na.rm = TRUE)
  ) |> 
  ungroup()
```

```{r s_stats_long}
# adjoin conditional son statistics to full data set
s_stats_long <- father_son_ht |> 
  left_join(
    y  = s_stats_per_f_mpt, 
    by = c("f_ivl", "f_mpt")
  )
```

```{r g_f_mpt}
g_f_mpt <- father_son_ht |> 
  filter(! is.na(f_mpt)) |> 
  ggplot(mapping = aes(
    x = f_mpt |> as_factor(), 
    y = son
  )) + 
  geom_boxplot()
g_f_mpt
```

```{r s_stats_per_f_mpt__show}
s_stats_per_f_mpt |> print(digits = 1)
```

The last column in the table above is the sample average of the son's height given the father's height, which we take as an estimate of the population average of the son's height given the father's height, that is the *conditional expectation* of son's height given father's height, as shown in the figure below.

```{r g_s_avg_per_f_mpt}
g_s_avg_per_f_mpt <- s_stats_per_f_mpt |> ggplot(mapping = aes(
  x = f_mpt, 
  y = s_avg, 
)) + 
  geom_line() + 
  geom_count(
    shape = 23, 
    size = s_stats_per_f_mpt$s_count |> sqrt())
g_s_avg_per_f_mpt
```

Galton noted that in general sons were taller than fathers, with extremely tall or short fathers corresponding to not quite so extremely tall or short sons, respectively. This observation led Galton to coin the term "regression toward mediocrity", or (more politely) "regression toward the mean".

## Class Exercise: Diamond Data

Team up with a classmate and load the diamond data provided by R package ggplot2. How many rows of data are there? How many columns? What questions occur to you about the data? How might you address those questions? Take 15 minutes to prepare to report out to the class.

```{r diamonds}
# make your own copy of the diamond data
diamonds <- ggplot2::diamonds
```

## Class Exercise: General Social Survey Data

The [General Social Survey](https://gss.norc.org/) is a long-running US survey conducted by the independent research organization NORC at the University of Chicago. The survey consists of thousands of questions designed to monitor changes in social characteristics and attitudes in the US. In R package forcats, Hadley Wickham formed the data set gss_cat consisting of just a few columns to illustrate the challenges one encounters when working with categorical (non-numeric) variables ("factors").

Team up with a classmate and load the gss_cat data-set provided by R package forcats. How many rows of data are there? How many columns? What questions occur to you about the data? How might you address those questions? Take 15 minutes to prepare to report out to the class.

```{r gss_cat}
# make your own copy of the survey data 
gss_cat <- forcats::gss_cat
```

## Discussion: what should EDA mean?

Exploratory Data Analysis (EDA) is an approach to data analysis advocated by [John Tukey](https://en.wikipedia.org/wiki/John_Tukey), a leading American statistician of the 20th century. The approach contrasts with what Tukey called "confirmatory analysis", that is, a focus on probability models of data-generation along with the estimation or testing of model parameters. The difference is one of emphasis: EDA includes models suggested by data, but with an emphasis on understanding current and potential data sets.

The exploration is led by one's questions about the data. Relevant questions may or may not be obvious (or given). Variables may or may not be readily categorized as "response variables" versus "predictor variables". The ability to develop and recognize relevant questions is an important skill largely gained through experience.

Important EDA outcomes include

-   the discovery of unanticipated data patterns, and

-   proposals to examine tentative answers suggested by the current data, perhaps using a new data set designed for this purpose.

EDA methods are used within the context of confirmatory analysis to examine the data for errors not encompassed by the models under study (e.g., errors in data transcription or transmission), or to search for other departures from model assumptions.

EDA methods can be broadly understood as the methods of descriptive statistics: data summaries (graphical or tabular) intended to enhance our understanding of the data. EDA differs from descriptive statistics in a reliance on the questions of the data analyst and a readiness to examine various transformations of the data.

As an example, here are some ways we might address the question of how, if at all, the heights of fathers and sons differ in the data presented above.

```{r height_per_fs}
# render "father" and "son" as levels of factor "fs"
height_per_fs <- father_son_ht |> 
  # continue to identify father-son pairs
  mutate(idx = 1:nrow(father_son_ht)) |>
  dplyr::select(idx, father, son) |> 
  pivot_longer(
    cols = c(father, son), 
    names_to = "fs", 
    values_to = "height"
  )
```

```{r g_ht_per_fs}
g_ht_per_fs <- height_per_fs |> 
  ggplot(mapping = aes(x = fs |> as_factor(), y = height)) + 
  geom_boxplot()
g_ht_per_fs
```

```{r ht_per_fs__summarise}
fs_smy <- height_per_fs |> dplyr::select(fs, height) |>
  group_by(fs) |> 
  summarise(
    min = min(height), 
    mid = median(height), 
    avg = mean(height), 
    max = max(height)
  )
fs_smy |> print(digits = 1)
```

The box-plot and table above show that, on average, sons are about an inch taller than fathers. Here's a histogram of the difference in heights (son minus father) across father-son pairs.

```{r s_minus_f}
father_son_ht <- father_son_ht |> 
  mutate(s_minus_f = son - father)
```

```{r s_minus_f__hist}
g_s_minus_f <- father_son_ht |> 
  ggplot(mapping = aes(x = s_minus_f)) + 
  geom_histogram()
g_s_minus_f
```

```{r s_minus_f__smy}
father_son_ht$s_minus_f |> 
  summary() |> 
  print(digits = 1)
```

The summary of individual differences in height (son minus father) strengthens the previous aggregate summaries: the distribution of son's height minus father's height is fairly symmetric around a difference of about one inch.

## Team Exercises

1.  Response versus predictor variables: for each of the data sets presented above, propose one or more variables as response variables. How, if at all, might someone else argue for a different choice? What should we mean by "response" and "predictor" variables? Describe a situation in which this distinction would not be suitable.

2.  Regression to the mean: the discussion of father-son heights asserts that: "extremely tall or short fathers \[correspond\] to not quite so extremely tall or short sons, respectively". How would you formulate the meaning of that phrase? Do the data demonstrate this phenomenon?

3.  Diamond data: propose a question for the diamond data, and then try to address that question. What (if anything) did you learn from this task?

4.  Survey data: propose a question for the gss_cat data, and then try to address that question. What (if anything) did you learn from this task?

## Resources

[R Graphics Cookbook (2e)](https://r-graphics.org/) by Winston Chang
